---
title: "Mechanistic Modeling"
author: "Olofin Daniel"
date: "2025-07-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

 # Mechanistic Modeling - Approach 1:

## Theoretical Framework

Mechanistic modeling provides a framework to simulate chromosomal instability (CIN) based on biological assumptions. Instead of testing statistical hypotheses, we aim to replicate the **stochastic behavior** of cells exposed to different treatments using **probabilistic transition rules**.

In this model, each **chromosome** in a **single cell clone** can either remain the same, gain a copy, or lose a copy at each time step. These changes depend on the treatment condition, reflecting biological stress-induced instability (e.g., oxidative stress, nutrient deprivation, mitotic spindle disruptions).

### Probabilistic Transition Rules
Each chromosome in each cell may either:
- Gain a copy (+1),
- Lose a copy (−1), or
- Remain unchanged (0),
at each time step.

Let:
- \( C_{i,t} \): copy number of chromosome \( i \) in a given cell at time \( t \),
- \( P^T_g \): probability of chromosome gain under treatment \( T \),
- \( P^T_l \): probability of chromosome loss under treatment \( T \).

**Transition Probabilities:**
\[
C_{i,t+1} = \begin{cases} 
C_{i,t} + 1 & \text{with probability } P^T_g \\
C_{i,t} - 1 & \text{with probability } P^T_l \text{ (if } C_{i,t} > 0) \\
C_{i,t}     & \text{with probability } 1 - P^T_g - P^T_l
\end{cases}
\]


The probabilities are treatment-specific:

```{r}
# Load necessary libraries
library(tidyverse)
library(reshape2)
library(pheatmap)

# Define simulation parameters
set.seed(123)

n_cells <- 1000
n_chromosomes <- 22
n_time <- 5

# Define treatments and transition probabilities
treatments <- c("Control", "Glucose", "Phosphate", 
                "Reversine_Control", "Reversine_Glucose", 
                "RPMI", "K_Condition", "Oxygen", "R_Condition")

transition_probs <- list(
  Control = list(gain = 0.01, loss = 0.01),
  Glucose = list(gain = 0.10, loss = 0.02),
  Phosphate = list(gain = 0.06, loss = 0.04),
  Reversine_Control = list(gain = 0.08, loss = 0.06),
  Reversine_Glucose = list(gain = 0.15, loss = 0.12),
  RPMI = list(gain = 0.05, loss = 0.03),
  K_Condition = list(gain = 0.07, loss = 0.02),
  Oxygen = list(gain = 0.03, loss = 0.01),
  R_Condition = list(gain = 0.11, loss = 0.05)
)

chromosomes <- paste0("V", 1:n_chromosomes)

# Simulation function
simulate_karyotypes <- function(treatment, n_cells, n_chromosomes, n_time) {
  gain_p <- transition_probs[[treatment]]$gain
  loss_p <- transition_probs[[treatment]]$loss
  
  mat <- matrix(2, nrow = n_cells, ncol = n_chromosomes)
  colnames(mat) <- chromosomes
  
  for (t in 1:n_time) {
    for (i in 1:n_cells) {
      for (j in 1:n_chromosomes) {
        r <- runif(1)
        if (r < gain_p) {
          mat[i, j] <- mat[i, j] + 1
        } else if (r < gain_p + loss_p && mat[i, j] > 0) {
          mat[i, j] <- mat[i, j] - 1
        }
      }
    }
  }
  
  df <- as.data.frame(mat)
  df$treatment <- treatment
  df$cell_id <- 1:nrow(df)
  return(df)
}

# Run simulation for all treatments
simulated_data <- map_dfr(treatments, ~simulate_karyotypes(.x, n_cells, n_chromosomes, n_time))

# Summarize and plot heatmap
mean_profiles <- simulated_data %>%
  pivot_longer(cols = all_of(chromosomes), names_to = "chromosome", values_to = "count") %>%
  group_by(treatment, chromosome) %>%
  summarise(mean_count = mean(count), .groups = "drop") %>%
  pivot_wider(names_from = chromosome, values_from = mean_count)

heat_matrix <- as.matrix(mean_profiles[,-1])
rownames(heat_matrix) <- mean_profiles$treatment

# Visualize
pheatmap(heat_matrix, cluster_rows = TRUE, cluster_cols = TRUE,
         main = "Avg Chromosome Count per Treatment",
         color = colorRampPalette(c("blue", "white", "red"))(100))

```


# Appoarch 2:

## Theoretical Framework

We model karyotype evolution as a discrete-time Markov process.
Each chromosome copy number evolves independently in each cell according to gain/loss probabilities.

\[
P(C_{i,t+1} = c' | C_{i,t} = c) =
\begin{cases}
P_g^T & c' = c + 1\\
P_l^T & c' = \max(0, c - 1)\\
1 - P_g^T - P_l^T & c' = c
\end{cases}
\]

Where:
- \( P_g^T \) is the probability of gain under treatment \( T \)
- \( P_l^T \) is the probability of loss under treatment \( T \)

### Transition matrix for a single chromosome (states 0 to 5 for illustration)

\[
\mathbf{P}^T =
\begin{bmatrix}
1 - P_g^T & P_g^T & 0 & 0 & 0 & 0 \\
P_l^T & 1 - P_g^T - P_l^T & P_g^T & 0 & 0 & 0 \\
0 & P_l^T & 1 - P_g^T - P_l^T & P_g^T & 0 & 0 \\
0 & 0 & P_l^T & 1 - P_g^T - P_l^T & P_g^T & 0 \\
0 & 0 & 0 & P_l^T & 1 - P_g^T - P_l^T & P_g^T \\
0 & 0 & 0 & 0 & P_l^T & 1 - P_l^T
\end{bmatrix}
\]

### Joint transition

\[
P(\mathbf{C}_{t+1} | \mathbf{C}_t) = \prod_{i=1}^{22} P(C_{i,t+1} | C_{i,t})
\]

# Simulation Data Structure

| clone_id  | replicate_id | line | treatment | X1 | X2 | ... | X22 |
|-----------|--------------|-------|------------|----|----|------|------|
| CLONE_1 | SNU-668_G1_A10_seed | SNU-668 | Glucose | 2 | 3 | ... | 1 |
| CLONE_2 | SNU-668_G1_A10_seed | SNU-668 | Glucose | 1 | 2 | ... | 2 |
| ... | ... | ... | ... | ... | ... | ... | ... |
| CLONE_1000 | SNU-668_G1_A10_seed | SNU-668 | Glucose | 2 | 2 | ... | 3 |

This structure enables both single-cell and replicate-level analysis.

# Flowchart

```{r}
library(dplyr)
library(tidyr)
library(vegan)
library(ggplot2)
library(DiagrammeR)

# Define treatments and their gain/loss probabilities (can be tuned)
treatment_params <- data.frame(
  treatment = c("Control", "Glucose", "Phosphate", "Reversine_Control", 
                "Reversine_Glucose", "RPMI", "K_Condition", "Oxygen", "R_Condition"),
  P_gain = c(0.02, 0.03, 0.03, 0.05, 0.06, 0.02, 0.04, 0.03, 0.04),
  P_loss = c(0.015, 0.02, 0.02, 0.04, 0.05, 0.015, 0.03, 0.02, 0.03)
)

cell_lines <- c("HGC-27", "MDAMB231", "SNU-668", "SUM-159", "SUM-159-4N")

# Example replicate IDs (subset for demo)
replicate_id <- paste0(rep(cell_lines, each=2), "_Rep", 1:2)

# Simulate karyotypes
generate_karyotypes <- function(n_cells, n_chrom = 22, time_steps = 10, P_gain, P_loss) {
  karyo_list <- vector("list", n_cells)
  for (cell in 1:n_cells) {
    karyo <- rep(2, n_chrom)
    for (t in 1:time_steps) {
      change <- sample(c(+1, -1, 0), n_chrom, replace=TRUE,
                       prob=c(P_gain, P_loss, 1 - P_gain - P_loss))
      karyo <- pmax(0, karyo + change) # no negative copy numbers
    }
    karyo_list[[cell]] <- karyo
  }
  do.call(rbind, karyo_list)
}

set.seed(123)
sim_data <- list()
clone_counter <- 1
for (rep in replicate_id) {
  line <- strsplit(rep, "_")[[1]][1]
  treatment <- sample(treatment_params$treatment, 1)
  P_g <- treatment_params$P_gain[treatment_params$treatment == treatment]
  P_l <- treatment_params$P_loss[treatment_params$treatment == treatment]
  karyos <- generate_karyotypes(n_cells = 1000, P_gain = P_g, P_loss = P_l)
  sim_data[[rep]] <- data.frame(
    clone_id = paste0("CLONE_", clone_counter:(clone_counter + 999)),
    replicate_id = rep,
    line = line,
    treatment = treatment,
    karyos
  )
  clone_counter <- clone_counter + 1000
}
sim_df <- bind_rows(sim_data)

# Aggregate to replicate centroids
karyo_df <- sim_df %>%
  group_by(replicate_id, treatment, line) %>%
  summarise(across(starts_with("X"), mean), .groups = "drop")

# Create matrix of karyotype values
karyo_mat <- as.matrix(karyo_df[, grep("^X", names(karyo_df))])
rownames(karyo_mat) <- karyo_df$replicate_id

# Create metadata
meta_flask <- karyo_df[, c("replicate_id", "line", "treatment")]
rownames(meta_flask) <- karyo_df$replicate_id

# Compute distance matrix
dist_flask <- dist(karyo_mat)

# Check alignment
stopifnot(all(rownames(meta_flask) == attr(dist_flask, "Labels")))


# --- Flowchart ---
grViz(" 
digraph karyo_model {
  graph [rankdir = TB]
  node [shape = rectangle, style = filled, fillcolor = lightblue]
  Start [label = 'Start: Diploid karyotype']
  TimeStep [label = 'For each time step']
  ChromLoop [label = 'For each chromosome']
  GainLoss [label = 'Apply gain/loss probabilities']
  Update [label = 'Update karyotype (no negatives)']
  Output [label = 'Record final karyotype']
  
  Start -> TimeStep -> ChromLoop -> GainLoss -> Update -> TimeStep
  TimeStep -> Output
}")

# Visualization (optional)
karyo_long <- sim_df %>% pivot_longer(cols = starts_with("X"), names_to = "chrom", values_to = "copy")
ggplot(karyo_long, aes(copy)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~treatment) +
  theme_minimal() +
  labs(title = "Simulated Chromosome Copy Number Distributions")

```


```{r}
# Treatment probabilities as per your reference
treatment_params <- data.frame(
  treatment = c("Control", "Glucose", "Phosphate", "Reversine_Control", 
                "Reversine_Glucose", "RPMI", "K_Condition", "Oxygen", "R_Condition"),
  P_gain = c(0.02, 0.03, 0.03, 0.05, 0.06, 0.02, 0.04, 0.03, 0.04),
  P_loss = c(0.015, 0.02, 0.02, 0.04, 0.05, 0.015, 0.03, 0.02, 0.03)
)

# Function to create matrix for one treatment
create_transition_matrix <- function(P_gain, P_loss) {
  P_same <- 1 - P_gain - P_loss
  mat <- matrix(0, nrow = 4, ncol = 4)  # For 0,1,2,3 copy numbers
  
  mat[1,1] <- P_same + P_loss
  mat[1,2] <- P_gain
  
  mat[2,1] <- P_loss
  mat[2,2] <- P_same
  mat[2,3] <- P_gain
  
  mat[3,2] <- P_loss
  mat[3,3] <- P_same
  mat[3,4] <- P_gain
  
  mat[4,3] <- P_loss
  mat[4,4] <- P_same + P_gain
  
  return(mat)
}

# Apply to all treatments
transition_matrices <- lapply(1:nrow(treatment_params), function(i) {
  row <- treatment_params[i, ]
  mat <- create_transition_matrix(row$P_gain, row$P_loss)
  list(treatment = row$treatment, matrix = mat)
})

# Display nicely
for (item in transition_matrices) {
  cat("\n--- Transition Matrix for", item$treatment, "---\n")
  print(round(item$matrix, 3))
}

```



```{r}
# Compute average copy number per treatment and chromosome
heat_matrix <- sim_df %>%
  pivot_longer(cols = starts_with("X"), names_to = "chrom", values_to = "copy") %>%
  group_by(treatment, chrom) %>%
  summarise(mean_copy = mean(copy), .groups = 'drop') %>%
  pivot_wider(names_from = chrom, values_from = mean_copy) %>%
  column_to_rownames("treatment") %>%
  as.matrix()

# Plot heatmap
pheatmap(heat_matrix, 
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         main = "Avg Chromosome Count per Treatment",
         color = colorRampPalette(c("blue", "white", "red"))(100))
```

**Note: I tried to run the PERMANOVA using the appoarch two data, the result seems different from the actaul anaylsis but supported the first result I got when Collapsing the cell-level karyotypes into flask-level centroids:**

```{r}
# Run PERMANOVA
library(vegan)
adonis2(
  formula = dist_flask ~ line + treatment,
  data = meta_flask,
  strata = meta_flask$line,
  permutations = 999
)

```
A PERMANOVA was conducted to assess whether differences in karyotype profiles among individual cells were significantly associated with cell line and treatment conditions, while accounting for stratified permutation within each cell line. The overall model was not significant ($F$ = 3.39, $p$ = 1.000), despite explaining approximately 96.4% of the total variation in the chromosome features ($R^2$ = 0.964).






**Equal Probability**

```{r}
library(dplyr)
library(tidyr)
library(vegan)
library(ggplot2)
library(DiagrammeR)

# Define treatments and their gain/loss probabilities (can be tuned)
treatment_params <- data.frame(
  treatment = c("Control", "Glucose", "Phosphate", "Reversine_Control", 
                "Reversine_Glucose", "RPMI", "K_Condition", "Oxygen", "R_Condition"),
  P_gain = 0.02,
  P_loss = 0.02
)

cell_lines <- c("HGC-27", "MDAMB231", "SNU-668", "SUM-159", "SUM-159-4N")

# Example replicate IDs (subset for demo)
replicate_id <- paste0(rep(cell_lines, each=2), "_Rep", 1:2)

# Simulate karyotypes
generate_karyotypes <- function(n_cells, n_chrom = 22, time_steps = 10, P_gain, P_loss) {
  karyo_list <- vector("list", n_cells)
  for (cell in 1:n_cells) {
    karyo <- rep(2, n_chrom)
    for (t in 1:time_steps) {
      change <- sample(c(+1, -1, 0), n_chrom, replace=TRUE,
                       prob=c(P_gain, P_loss, 1 - P_gain - P_loss))
      karyo <- pmax(0, karyo + change) # no negative copy numbers
    }
    karyo_list[[cell]] <- karyo
  }
  do.call(rbind, karyo_list)
}

set.seed(123)
sim_data <- list()
clone_counter <- 1
for (rep in replicate_id) {
  line <- strsplit(rep, "_")[[1]][1]
  treatment <- sample(treatment_params$treatment, 1)
  P_g <- treatment_params$P_gain[treatment_params$treatment == treatment]
  P_l <- treatment_params$P_loss[treatment_params$treatment == treatment]
  karyos <- generate_karyotypes(n_cells = 1000, P_gain = P_g, P_loss = P_l)
  sim_data[[rep]] <- data.frame(
    clone_id = paste0("CLONE_", clone_counter:(clone_counter + 999)),
    replicate_id = rep,
    line = line,
    treatment = treatment,
    karyos
  )
  clone_counter <- clone_counter + 1000
}
sim_df <- bind_rows(sim_data)

# Aggregate to replicate centroids
karyo_df <- sim_df %>%
  group_by(replicate_id, treatment, line) %>%
  summarise(across(starts_with("X"), mean), .groups = "drop")

# Create matrix of karyotype values
karyo_mat <- as.matrix(karyo_df[, grep("^X", names(karyo_df))])
rownames(karyo_mat) <- karyo_df$replicate_id

# Create metadata
meta_flask <- karyo_df[, c("replicate_id", "line", "treatment")]
rownames(meta_flask) <- karyo_df$replicate_id

# Compute distance matrix
dist_flask <- dist(karyo_mat)

# Check alignment
stopifnot(all(rownames(meta_flask) == attr(dist_flask, "Labels")))


# --- Flowchart ---
grViz(" 
digraph karyo_model {
  graph [rankdir = TB]
  node [shape = rectangle, style = filled, fillcolor = lightblue]
  Start [label = 'Start: Diploid karyotype']
  TimeStep [label = 'For each time step']
  ChromLoop [label = 'For each chromosome']
  GainLoss [label = 'Apply gain/loss probabilities']
  Update [label = 'Update karyotype (no negatives)']
  Output [label = 'Record final karyotype']
  
  Start -> TimeStep -> ChromLoop -> GainLoss -> Update -> TimeStep
  TimeStep -> Output
}")

# Visualization (optional)
karyo_long <- sim_df %>% pivot_longer(cols = starts_with("X"), names_to = "chrom", values_to = "copy")
ggplot(karyo_long, aes(copy)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~treatment) +
  theme_minimal() +
  labs(title = "Simulated Chromosome Copy Number Distributions")
# Compute average copy number per treatment and chromosome
heat_matrix <- sim_df %>%
  pivot_longer(cols = starts_with("X"), names_to = "chrom", values_to = "copy") %>%
  group_by(treatment, chrom) %>%
  summarise(mean_copy = mean(copy), .groups = 'drop') %>%
  pivot_wider(names_from = chrom, values_from = mean_copy) %>%
  column_to_rownames("treatment") %>%
  as.matrix()

# Plot heatmap
pheatmap(heat_matrix, 
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         main = "Avg Chromosome Count per Treatment",
         color = colorRampPalette(c("blue", "white", "red"))(100))

```

# Theoretical Framework

## Objective

This simulation models the chromosomal instability (CIN) in cancer cells by simulating the copy number evolution of chromosomes across time under different treatments and cell lines. The aim is to generate realistic karyotype variability expected in tumors without pre-specified hypotheses about treatment effects.

## Assumptions

- The initial copy number of each chromosome is randomly drawn from a uniform distribution between 1 and 5, reflecting the inherent aneuploidy of cancer cells:
\[
C_{i,0} \sim \text{Uniform}(1, 5)
\]

- Chromosomal changes occur at discrete time steps. At each time step, each chromosome may:
  - **Gain a copy**: \(C_{i,t+1} = C_{i,t} + 1\)
  - **Lose a copy** (if \(C_{i,t} > 0\)): \(C_{i,t+1} = C_{i,t} - 1\)
  - **Remain unchanged**: \(C_{i,t+1} = C_{i,t}\)

- Probabilities of gain and loss are equal across all treatments:
\[
P_{\text{gain}} = P_{\text{loss}} = p
\]
where \(p = 0.05\).

- The probability of no change is:
\[
P_{\text{no change}} = 1 - 2p
\]

## Transition Probability Matrix

At each time step for each chromosome:
\[
P(C_{i,t+1}|C_{i,t}) =
\begin{bmatrix}
P_{\text{gain}} & P_{\text{loss}} & P_{\text{no change}}
\end{bmatrix}
=
\begin{bmatrix}
0.05 & 0.05 & 0.90
\end{bmatrix}
\]

```{r}
library(dplyr)
library(tidyr)

# Setup
set.seed(123)
n_cells <- 1000
n_chrom <- 22
n_steps <- 10
p_gain <- 0.05
p_loss <- 0.05

treatments <- c("Control", "Glucose", "Phosphate", "Reversine_Control", 
                "Reversine_Glucose", "RPMI", "K_Condition", "Oxygen", "R_Condition")

cell_lines <- c("HGC-27", "MDAMB231", "SNU-668", "SUM-159", "SUM-159-4N")

chromosomes <- paste0("V", 1:n_chrom)

# Function
simulate_cancer_karyotypes <- function(n_cells, n_chrom, n_steps, p_gain, p_loss, treatment, line) {
  karyo_list <- vector("list", n_cells)
  for (cell in 1:n_cells) {
    karyo <- sample(1:5, n_chrom, replace = TRUE)
    names(karyo) <- chromosomes
    
    for (t in 1:n_steps) {
      change <- sapply(karyo, function(count) {
        move <- sample(c(1, -1, 0), 1,
                       prob = c(p_gain, ifelse(count > 0, p_loss, 0), 
                                1 - p_gain - ifelse(count > 0, p_loss, 0)))
        return(move)
      })
      karyo <- pmax(0, karyo + change)
    }
    karyo_list[[cell]] <- data.frame(t(karyo), cell_id = paste0("Cell_", cell),
                                     treatment = treatment, line = line)
  }
  return(bind_rows(karyo_list))
}

# Simulate for all cell lines and treatments
sim_results <- list()
counter <- 1
for (line in cell_lines) {
  for (treat in treatments) {
    sim_results[[counter]] <- simulate_cancer_karyotypes(n_cells = 100, # adjust as needed
                                                         n_chrom = n_chrom,
                                                         n_steps = n_steps,
                                                         p_gain = p_gain,
                                                         p_loss = p_loss,
                                                         treatment = treat,
                                                         line = line)
    counter <- counter + 1
  }
}

sim_df <- bind_rows(sim_results)

# View part of the result
head(sim_df)


# Long format for histogram
karyo_long <- sim_df %>% 
  pivot_longer(cols = starts_with("X"), names_to = "chrom", values_to = "copy")

# Histogram
ggplot(karyo_long, aes(copy)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  facet_wrap(~treatment) +
  theme_minimal() +
  labs(title = "Simulated Chromosome Copy Number Distributions")

# Compute mean copy number per treatment × chromosome
heat_data <- karyo_long %>%
  group_by(treatment, chrom) %>%
  summarise(mean_copy = mean(copy), .groups = 'drop') %>%
  pivot_wider(names_from = chrom, values_from = mean_copy)

# Convert to matrix
heat_matrix <- heat_data %>%
  column_to_rownames("treatment") %>%
  as.matrix()

# Z-score normalize (column-wise: across treatments per chromosome)
heat_matrix_z <- scale(heat_matrix, center = TRUE, scale = TRUE)

# Heatmap
pheatmap(heat_matrix_z,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         main = "Z-score Normalized Avg Chromosome Count per Treatment",
         color = colorRampPalette(c("blue", "white", "red"))(100))

```