---
title: "Permanova new"
author: "Olofin Daniel"
date: "2025-06-17"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, fig.width = 7, fig.height = 5)
```

```{r eval=FALSE, include=FALSE}
##Noemi has added you to the cloneid database, believe danielolofin is your username 
## and Karyotyping_Analysis is the password. Access is only possible when you are on the Moffitt network.
## If you email Noemi your home IP and/or moffitt guest IP she can give you access on those networks too.

#create user 'danielolofin'@'206.81.166.156' identified by 'Karyotyping_Analysis';
#GRANT SELECT ON *.* TO 'danielolofin'@'206.81.166.156';  
#create user 'danielolofin'@'192.168.1.32' identified by 'Karyotyping_Analysis';
#GRANT SELECT ON *.* TO 'danielolofin'@'192.168.1.32';  

```

```{r include=FALSE}
# Libraries
library(RMariaDB)
library(DBI)
library(tidyverse)
library(dplyr)  
library(pheatmap)
library(ggplot2)
library(reshape2)
library(lme4)     # for linear mixed models
library(lmerTest) # for p-values in lmer
library(emmeans)  # for posthoc comparisons
library(ggflowchart)
library(car)
library(sjPlot)
library(webshot)
library(stargazer)
library(factoextra)
library(FactoMineR)
library(vegan)

```

```{r include=FALSE}

db <- dbConnect(MariaDB(), dbname="CLONEID", 
                host="cloneid.cswgogbb5ufg.us-east-1.rds.amazonaws.com", 
                user="danielolofin",
                  password="Karyotyping_Analysis")

query <- "SELECT id, passaged_from_id1, passage, event, date, media from Passaging WHERE passaged_from_id1"
mismatches <- dbGetQuery(db, query)

```

```{r include=FALSE}
## this is the database containing all the data. I use R to work with it but if you have experience with SQL database interactions
## you can access with any other software you like
dbListTables(db)

```

```{r include=FALSE}
dbListFields(db,"Perspective") ##karyotype data is in here.
```

```{r include=FALSE}
dbListFields(db,"Passaging") ## a lot of the metadata is here
#> dbListFields(db,"Passaging")
```

```{r include=FALSE}
dbListFields(db,"Media") ## Media info, the id column here matches the media column inside passaging.#
```



```{r include=FALSE}
## this is how to pull the karyotype data from the perspective table. For this you'll need to know the origin.
## Perhaps vural can provide a list of all the origins that belong in the experiment. "Origin" maps to "id" field in the 
## passaging table.

q <- paste0("SELECT * FROM Perspective WHERE origin='SNU-668_G1_A10_seed' AND whichPerspective='GenomePerspective'")
rs <- dbSendQuery(db, q)
```

```{r include=FALSE}
# Single query to pull all relevant profiles
q <- paste0("SELECT * FROM Perspective WHERE origin='SNU-668_G1_A10_seed' AND whichPerspective='GenomePerspective'")
rs <- dbSendQuery(db, q)
res <- dbFetch(rs)
dbClearResult(rs)

# Convert each BLOB to a double vector
kvecs <- lapply(res$profile, function(p) {
  raw_vec <- p
  readBin(raw_vec, what = "double", n = length(raw_vec) / 8, endian = "big")
})

# Stack into a matrix
kmat <- do.call(rbind, kvecs)
rownames(kmat) <- res$cloneID
```


```{r include=FALSE}
# Define origins and build query
origins <- c(
  "SNU-668_R1_A55_seedT1", "SNU-668_R3_A55_seedT1", "SNU-668_K1_A34_seedT1",
  "SNU-668_K2_A34_seedT1", "SNU-668_K3_A34_seedT1", "HGC-27_RevG_A9-F_seed",
  "HGC-27_RevG2_A7-F_seed", "SNU-668_G1_A4_seed", "SNU-668_G2_A4_seed",
  "HGC-27_RevC_A45_seed", "SNU-668_RevG_A3_seed", "SNU-668_P2_A18k_seed",
  "SNU-668_P0_A11K_seed", "SNU-668_r2_A55_seedT1", "SNU-668_G1_A10_seed",
  "SNU-668_G2_A10_seed", "SNU-668_RevG_A10_seed", "HGC-27_G1_A9-F_seed",
  "HGC-27_G2_A9_harvesT155", "HGC-27_G2_A9-F_seed", "HGC-27_G1_A9_harvesT155",
  "HGC-27_RevG_A9_harvesT149", "HGC-27_RevG2_A8_seed", "MDA-MB-231_P_A2_harvest",
  "HGC-27_C_A45_seed", "SNU-668_RevC_A27_seed", "SNU-668_C_A24_seed",
  "SNU-668_C_A4_seed", "SNU-668_RevC_A3_seed", "MDA-MB-231_RPMI_A12_VT2_seed",
  "SNU-668_P1_A18K_seed", "SNU-668_O1_A3_exp_seedT4", "SNU-668_O2_A3_exp_seedT4",
  "HGC-27_P1_A18K_harvest", "HGC-27_P2_A18K_harvest", "SUM159_NLS_4N_O1_A7K_harvest",
  "SUM159_NLS_2N_O1_A7K_harvest", "SUM-159_NLS_4N_A5M_K_harvest",
  "SUM159_NLS_4N_O2_A7K_harvest", "SUM159_NLS_2N_O2_A7K_harvest",
  "SUM-159_NLS_2N_A7M_K_harvest"
)


## Passaging table
q <- paste0("SELECT * FROM Passaging")
rs <- dbSendQuery(db, q)
res_passage <- dbFetch(rs)
dbClearResult(rs)

names(res_passage)[1] <- "replicate_id"


origin_clause <- paste0("'", origins, "'", collapse = ", ")
 q <- paste0("SELECT * FROM Perspective WHERE origin IN (", origin_clause, ") AND whichPerspective='GenomePerspective'")



# Fetch data from the database
if (!dbIsValid(db)) {
  db <- dbConnect(...)  
}
rs <- dbSendQuery(db, q)
res <- dbFetch(rs)
dbClearResult(rs)

#  Decode the BLOBs into double vectors
kvecs <- lapply(res$profile, function(p) {
  raw_vec <- p
  readBin(raw_vec, what = "double", n = length(raw_vec) / 8, endian = "big")
})

```


# Creating a new cell line for SUM to seperate (2N and 4N)

```{r include=FALSE}
# Building the matrix
kmat <- do.call(rbind, kvecs)
rownames(kmat) <- res$cloneID



res$treatment_group <- case_when(
  grepl("_G1|_G2", res$origin) ~ "Glucose",
  grepl("_K1|_K2|_K3", res$origin) ~ "K_Condition",
  grepl("_P1|_P2|_P0|_P", res$origin) ~ "Phosphate",
  grepl("_R1|_R2|_R3|_r2", res$origin) ~ "R_Condition",
  grepl("_O1|_O2", res$origin) ~ "Oxygen",
  grepl("_RevC", res$origin) ~ "Reversine_Control",
  grepl("_RevG|_RevG2", res$origin) ~ "Reversine_Glucose",
  grepl("_C", res$origin) ~ "Control",
  grepl("_RPMI", res$origin) ~ "Control",
  grepl("_K", res$origin) ~ "Control",
  TRUE ~ NA_character_
)


# Seperating SNU_168 by number of their chomosome
res %>% 
    mutate(sampleSource = ifelse(str_detect(alias, "N4") & sampleSource == "SUM-159",
                               "SUM-159-4N",
                               sampleSource)) -> res




# Combining matrix and group metadata
karyo_df <- as.data.frame(kmat)
karyo_df$treatment <- res$treatment
karyo_df$cloneID <- res$cloneID
karyo_df$line<- res$sampleSource
karyo_df$replicate_id<- res$origin

unique(karyo_df$replicate_id)

karyo_df<- karyo_df %>%
  mutate(time_category = case_when(
    grepl("A([0-9]+)", replicate_id) & as.numeric(sub(".*_A([0-9]+).*", "\\1", replicate_id)) <= 10 ~ "Early",
    grepl("harvesT", replicate_id, ignore.case = TRUE) ~ "Late",
    grepl("A([0-9]+)", replicate_id) & as.numeric(sub(".*_A([0-9]+).*", "\\1", replicate_id)) >= 18 ~ "Late",
    TRUE ~ "Ambiguous"
  ))

```


# Grouping similar treatment conditions into biologically meaningful categories:


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyr)
library(ggplot2)

# Convert to long format for ggplot2
karyo_long <- karyo_df %>%
  pivot_longer(cols = starts_with("V"),
               names_to = "chromosome",
               values_to = "count")

# Boxplot for all chromosomes, faceted
ggplot(karyo_long, aes(x = treatment, y = count, fill = treatment)) +
  geom_boxplot(outlier.shape = NA, show.legend = F) +
  facet_wrap(~chromosome, scales = "free_y", ncol = 6) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Chromosome Counts by Treatment",
       y = "Chromosome Count")

```

```{r echo=FALSE}
# Compute total chromosome per cell
karyo_df$total_chr <- rowSums(karyo_df[, paste0("V", 1:22)])

# Histogram of total chromosome count
ggplot(karyo_df, aes(x = total_chr, fill = treatment)) +
  geom_histogram(binwidth = 1, color = "black", position = "identity", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Total Chromosome Count per Cell",
       x = "Total Chromosomes",
       y = "Number of Cells")

```

```{r}
# Mean per chromosome per treatment
mean_chr <- karyo_long %>%
  group_by(treatment, chromosome) %>%
  summarise(mean_count = mean(count, na.rm = TRUE), .groups = "drop")

ggplot(mean_chr, aes(x = chromosome, y = mean_count, fill = treatment)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Mean Chromosome Count per Treatment",
       y = "Mean Count")

```

```{r}
library(pheatmap)

# Make a cell-by-chromosome matrix
karyo_mat <- as.matrix(karyo_df[, paste0("V", 1:22)])
rownames(karyo_mat) <- karyo_df$cloneID
# Build an annotation data frame for treatment
annotation_row <- data.frame(treatment = karyo_df$treatment)
rownames(annotation_row) <- karyo_df$cloneID


# Simple heatmap
pheatmap(karyo_mat[1:50, ], cluster_rows = TRUE, cluster_cols = TRUE, 
         main = "Chromosome Profiles (Subset of Cells)",
         show_rownames = FALSE)

library(pheatmap)

# Create matrix
karyo_mat <- as.matrix(karyo_df[, paste0("V", 1:22)])
rownames(karyo_mat) <- karyo_df$cloneID

# Create annotation
annotation_row <- data.frame(treatment = karyo_df$treatment)
rownames(annotation_row) <- karyo_df$cloneID

# Z-score normalize by chromosome (column)
karyo_mat_z <- scale(karyo_mat)

# Check dimensions
stopifnot(all(rownames(karyo_mat_z) == rownames(annotation_row)))

# Plot heatmap
pheatmap(karyo_mat_z,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         annotation_row = annotation_row,
         main = "Z-score Normalized Chromosome Profiles (All Cells)",
         show_rownames = FALSE,
         color = colorRampPalette(c("blue", "white", "red"))(100))



# Make matrix of chromosome counts
karyo_mat <- as.matrix(karyo_df[, paste0("V", 1:22)])
rownames(karyo_mat) <- karyo_df$cloneID

# Create annotation for treatments
annotation_row <- data.frame(treatment = karyo_df$treatment)
rownames(annotation_row) <- karyo_df$cloneID

# Sanity check alignment
stopifnot(all(rownames(karyo_mat) == rownames(annotation_row)))

# Plot heatmap without normalization
pheatmap(karyo_mat,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         annotation_row = annotation_row,
         main = "Raw Chromosome Copy Numbers (All Cells)",
         show_rownames = FALSE,
         color = colorRampPalette(c("blue", "white", "red"))(100))


```


- **Note:** Cell-Level PERMANOVA without Collapsing Replicates because:
1. I am studying how treatment affects individual cells' karyotypes.
2. Each row represents a unique cell, preserving biological variability within flasks.
3. I stratify by line, so that permutations are done within each cell line.
4. So that I can retain statistical power, because you are using all the data (not just averages).

```{r}
# Load necessary libraries
library(vegan)
library(dplyr)
library(ggplot2)

# Step 1: Make sure your karyo_df and meta_flask align
# Let's say rownames were not preserved, rebuild them together

# Create distance matrix from a clean matrix
karyo_mat <- as.matrix(karyo_df[ , 1:22])
rownames(karyo_mat) <- karyo_df$replicate_id   # Set replicate_id as rownames

# Create metadata with matching replicate_id
meta_flask <- karyo_df[, c("replicate_id", "line", "treatment","time_category")]
rownames(meta_flask) <- seq_len(nrow(meta_flask))
rownames(karyo_mat) <- seq_len(nrow(karyo_mat))

# Now subset meta_flask to match the distance matrix
meta_flask <- meta_flask[rownames(karyo_mat), ]

# Recreate distance matrix and run adonis2
dist_flask <- dist(karyo_mat)

# Check alignment
stopifnot(all(rownames(meta_flask) == attr(dist_flask, "Labels")))

# Run adonis2 and save result
adonis_res <- adonis2(
  formula = dist_flask ~ line + treatment,
  data = meta_flask,
  strata = meta_flask$line,
  permutations = 999
)

adonis_res <- adonis2(
  formula = dist_flask ~ line + treatment + time_category,
  data = meta_flask,
  strata = meta_flask$line,
  permutations = 999
)



# Convert to data frame for export
adonis_df <- as.data.frame(adonis_res)

# Clean row names as a column
adonis_df <- tibble::rownames_to_column(adonis_df, "Effect")

# Print or view
print(adonis_df)

```

**Interpretation:** A PERMANOVA was conducted to assess whether differences in karyotype profiles among individual cells were significantly associated with cell line and treatment conditions, while accounting for stratified permutation within each cell line. The overall model was highly significant (F = 109.94, $p$ = 0.001), indicating that the combined effects of cell line and treatment explain approximately 63.4% of the total variation in the chromosome features ($R^{2}$ = 0.634). This suggests that both the intrinsic characteristics of different cell lines and the external interventions from treatments contribute meaningfully to chromosomal differences observed in the data. The remaining 36.6% of the variation is unexplained by these variables and may reflect biological heterogeneity, noise, or other unmeasured factors.

```{r}
# Step 6: Test if treatment affects karyotype variability (dispersion)
library(vegan)

# Step 1: Compute dispersion
dispersion <- betadisper(dist_flask, group = meta_flask$treatment)

# Step 2: Run ANOVA
dispersion_anova <- anova(dispersion)

# Step 3: Convert to tidy table
dispersion_df <- broom::tidy(dispersion_anova)

# Optional: Clean row names if needed
dispersion_df <- tibble::rownames_to_column(dispersion_df, var = "Effect")

# View or print
print(dispersion_df)
```
**Note:** We discussed about using variance. This PERMISP is meant to check the asumption of PERMANOV and also Multivariate dispersion (spread) across groups. it test Overall karyotype variability across treatments.


**Intepretation:** The analysis of multivariate dispersions (PERMDISP) revealed a significant difference in dispersion among treatment groups (F = 10.76, p < 0.001), suggesting that variability within treatments is not uniform. This violates a key assumption of PERMANOVA, which presumes homogeneous dispersions. Consequently, although significant PERMANOVA results may indicate group differences, part of this difference could stem from heterogeneous group variances (dispersion), rather than true differences in centroid locations. Therefore, any inference about treatment effects on karyotype composition should be interpreted with caution and supported with additional visualization (e.g., NMDS/CAPscale plots) or models robust to dispersion differences, such as generalized linear models.


```{r}
permutest(dispersion, permutations = 999, pairwise = T ) ->t
t$pairwise %>% kableExtra::kable(digits = 4, caption = "Pairwise Permuted P-Values for Group Dispersion")
```

**Interpretation:**
The pairwise PERMDISP analysis revealed significant differences in dispersion (i.e., within-group variability) among several treatment groups, indicating that the assumption of homogenous variance required by PERMANOVA is violated in many comparisons. Notably, treatments such as Glucose, Oxygen, R_Condition, and Reversine_Glucose showed significant differences in dispersion when compared with multiple other treatments (p < 0.05). For example, Glucose differed significantly from K_Condition, Phosphate, and R_Condition, while R_Condition differed in dispersion from nearly all other treatments, suggesting it may be especially destabilizing to karyotypic structure. In contrast, treatments like Control vs Reversine_Glucose and Reversine_Control vs Reversine_Glucose did not differ significantly in their dispersion, indicating comparable within-group variability. These findings imply that differences observed in PERMANOVA results could, at least in part, be driven by differences in dispersion rather than shifts in group centroids alone.

```{r}
plot(dispersion, hull = FALSE, main = "Group Dispersion Plot")
boxplot(dispersion, main = "Boxplot of Multivariate Dispersion by Group")


```




-   2: GLMM for Per-Chromosome Changes

```{r}
library(lme4)

# Binary outcome: gain defined as >2 copies
karyo_df$gain_chr08 <- ifelse(karyo_df$V8 > 2, 1, 0)

# GLMM: treatment as fixed effect, replicate as random effect
model_chr8 <- glmer(gain_chr08 ~ treatment + (1 | replicate_id), 
                    data = karyo_df, family = binomial)

summary(model_chr8)


```
**Note:** why use chr08 (This is reference to  the fact that biological relevance, in many cancer gain of chr08 are commonly associated with tumor progression).
**Interpretation:** The generalized linear mixed model assessing the impact of treatment on chromosome 08 gain revealed that no treatment condition had a statistically significant effect compared to the control. The control group showed a significantly high baseline probability of chr08 gain. Variability among replicate flasks was substantial, supporting the inclusion of random effects in the model.

## P-Adjustment (e.g., Bonferroni and Benjamini-Hochberg)
```{r}
# Extract p-values from model summary
p_vals <- summary(model_chr8)$coefficients[-1, "Pr(>|z|)"]  # Exclude Intercept

# Adjust p-values (Bonferroni)
p_adj_bonf <- p.adjust(p_vals, method = "bonferroni")

# Adjust p-values (Benjamini-Hochberg)
p_adj_bh <- p.adjust(p_vals, method = "BH")

# Combine results
p_table <- data.frame(
  Treatment = names(p_vals),
  Raw_P = p_vals,
  Bonferroni = p_adj_bonf,
  BH = p_adj_bh
)

p_table %>% kableExtra::kable()

```
**Interpretation:** A generalized linear mixed model (GLMM) was used to assess whether different treatment conditions significantly influenced the likelihood of gain of chromosome 08 across replicate cell populations. None of the treatments showed a statistically significant effect, either before or after adjusting for multiple testing using Bonferroni or Benjamini-Hochberg (BH) procedures. The lowest raw p-value was observed for the Oxygen treatment (p = 0.158), but this did not remain significant after adjustment. These findings suggest that gain of chromosome 08 is not significantly associated with any specific treatment condition in this dataset, implying that observed variations are likely due to random chance or other uncontrolled biological factors.


1. Repeat the GLMM for other chromosomes (chr07, chr13, etc.)

```{r}
library(lme4)
library(dplyr)
library(tidyr)

# Example for chr07 and chr13
karyo_df <- karyo_df %>%
  mutate(
    gain_chr07 = ifelse(V7 > 2, 1, 0),
    gain_chr13 = ifelse(V13 > 2, 1, 0)
  )

# Fit GLMMs for chr07 and chr13
fit_chr07 <- glmer(gain_chr07 ~ treatment + (1 | replicate_id),
                   data = karyo_df, family = binomial)
fit_chr13 <- glmer(gain_chr13 ~ treatment + (1 | replicate_id),
                   data = karyo_df, family = binomial)

# Summarize models
summary(fit_chr07)
summary(fit_chr13)
```


```{r}
for (cell_line in unique(karyo_df$line)) {
  cat("\n\n### Cell Line:", cell_line, "###\n")
  line_df <- subset(karyo_df, line == cell_line)
  line_mat <- vegdist(as.matrix(line_df[, 1:22]), method = "manhattan")
  
  print(adonis2(line_mat ~ treatment, 
                data = line_df, 
                permutations = 999))
}



library(vegan)
library(pairwiseAdonis)

library(vegan)

library(vegan)
library(dplyr)

# Initialize a list to hold pairwise results for each cell line
all_pairwise_results <- list()

for (cell_line in unique(karyo_df$line)) {
  cat("\n\n### Cell Line:", cell_line, "###\n")
  
  # Subset data for the current cell line
  line_df <- subset(karyo_df, line == cell_line)
  treatments <- unique(line_df$treatment)
  
  # Compute overall distance matrix
  line_mat <- vegdist(as.matrix(line_df[, 1:22]), method = "manhattan")
  
  # Run overall PERMANOVA
  cat("Overall PERMANOVA:\n")
  print(adonis2(line_mat ~ treatment, data = line_df, permutations = 999))
  
  # Pairwise comparisons (manual)
  pairwise_results <- data.frame()
  
  for (i in 1:(length(treatments) - 1)) {
    for (j in (i + 1):length(treatments)) {
      t1 <- treatments[i]
      t2 <- treatments[j]
      
      # Subset for two treatment groups
      sub_df <- line_df[line_df$treatment %in% c(t1, t2), ]
      
      # Check for replicates
      if (sum(sub_df$treatment == t1) < 2 || sum(sub_df$treatment == t2) < 2) next
      
      # Compute distance
      sub_mat <- vegdist(as.matrix(sub_df[, 1:22]), method = "manhattan")
      
      # Run pairwise PERMANOVA
      result <- adonis2(sub_mat ~ treatment, data = sub_df, permutations = 999)
      
      # Collect results
      pairwise_results <- rbind(pairwise_results, data.frame(
        Cell_Line = cell_line,
        Group1 = t1,
        Group2 = t2,
        F_value = result$F[1],
        P_value = result$`Pr(>F)`[1]
      ))
    }
  }
  
  # Adjust p-values
  if (nrow(pairwise_results) > 0) {
    pairwise_results$P_adj_BH <- p.adjust(pairwise_results$P_value, method = "BH")
    pairwise_results$P_adj_Bonf <- p.adjust(pairwise_results$P_value, method = "bonferroni")
    all_pairwise_results[[cell_line]] <- pairwise_results
  } else {
    cat("Not enough replicates per group for pairwise comparisons.\n")
  }
}



```
**Interpretation:** PERMANOVA analysis within individual cell lines revealed that treatment significantly influenced karyotypic variation in HGC-27 ($R^{2}$ = 12.2%, $p$ = 0.001), MDAMB231 ($R^{2}$ = 19.7%, $p$ = 0.001), and SNU-668 ($R^{2}$ = 16.1%, $p$ = 0.001), suggesting strong treatment effects on chromosomal composition within these lines. In contrast, no significant treatment effects were observed in SUM-159 ($R^{2}$ = 0.07%, $p$ = 0.848) and SUM-159-4N ($R^{2}$ = 2.9%, $p$ = 0.229), indicating greater karyotypic stability or treatment resistance in those lines.

```{r}
# Optional: combine all into one table
combined_pairwise_results <- do.call(rbind, all_pairwise_results)

# View combined results
combined_pairwise_results %>% kableExtra::kable()


```

**Interpretation:** The pairwise PERMANOVA analysis across four cell lines (HGC-27, MDAMB231, SNU-668, SUM-159, and SUM-159-4N) revealed strong treatment-specific karyotypic divergence, particularly in HGC-27 and SNU-668, while minimal or no effects were observed in SUM-159 lines.

- HGC-27: All 10 pairwise comparisons were statistically significant, even after both Benjamini-Hochberg (BH) and Bonferroni corrections. This indicates robust and consistent differences in chromosomal profiles between treatments such as Control vs. Glucose, Phosphate, and both Reversine conditions. These results underscore HGC-27’s high karyotypic sensitivity to environmental or drug-induced perturbations.

- MDAMB231: Only one pairwise comparison (Phosphate vs. RPMI) was significant (p < 0.001), suggesting a more specific and limited treatment effect on karyotype composition in this line.

- SNU-668: This line showed extensive significant pairwise differences across treatments. 27 out of 28 pairs were significant after BH adjustment (many also Bonferroni corrected), especially comparisons involving K_Condition, Oxygen, Phosphate, and Reversine_Glucose. This makes SNU-668 the most karyotypically plastic line in response to diverse conditions.

- SUM-159 & SUM-159-4N: No pairwise comparisons reached significance. This suggests that karyotype profiles remain stable in these lines across the tested treatments, indicating low chromosomal variability or insensitivity to the perturbations.





```{r}
# Step 7: Visualize results using PCoA (ordination plot)
pcoa <- cmdscale(dist_flask, eig = TRUE, k = 2)
pcoa_df <- data.frame(PC1 = pcoa$points[,1],
                      PC2 = pcoa$points[,2],
                      treatment = meta_flask$treatment,
                      line = meta_flask$line)

ggplot(pcoa_df, aes(x = PC1, y = PC2, color = treatment, shape = line)) +
  geom_point(size = 4) +
  theme_minimal() +
  ggtitle("PCoA of Karyotype Profiles (Flask Centroids)")

```

```{r}
# MDS using Manhattan Distance

# Load necessary libraries
library(vegan)
library(ggplot2)
library(dplyr)


# Step 1: Extract the chromosome features (V1 to V22)
karyo_matrix <- as.matrix(karyo_df[, paste0("V", 1:22)])

# Step 2: Compute Manhattan distance
dist_manhattan <- dist(karyo_matrix, method = "manhattan")

# Step 3: Perform MDS
mds_fit <- cmdscale(dist_manhattan, k = 2, eig = TRUE)

# Step 4: Prepare metadata
mds_df <- as.data.frame(mds_fit$points)
colnames(mds_df) <- c("Dim1", "Dim2")
mds_df$replicate_id <- karyo_df$replicate_id
mds_df$treatment <- karyo_df$treatment
mds_df$line <- karyo_df$line

# Step 5: Visualize
ggplot(mds_df, aes(x = Dim1, y = Dim2, color = treatment, shape = line)) +
  geom_point(alpha = 0.7, size = 3) +
  labs(title = "MDS Plot using Manhattan Distance",
       x = "Dimension 1", y = "Dimension 2") +
  theme_minimal(base_size = 14)
```



