---
title: "Permanova"
author: "Olofin Daniel"
date: "2025-06-06"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, fig.width = 7, fig.height = 5)
```

```{r eval=FALSE, include=FALSE}
##Noemi has added you to the cloneid database, believe danielolofin is your username 
## and Karyotyping_Analysis is the password. Access is only possible when you are on the Moffitt network.
## If you email Noemi your home IP and/or moffitt guest IP she can give you access on those networks too.

#create user 'danielolofin'@'206.81.166.156' identified by 'Karyotyping_Analysis';
#GRANT SELECT ON *.* TO 'danielolofin'@'206.81.166.156';  
#create user 'danielolofin'@'192.168.1.32' identified by 'Karyotyping_Analysis';
#GRANT SELECT ON *.* TO 'danielolofin'@'192.168.1.32';  

```

```{r include=FALSE}
# Libraries
library(RMariaDB)
library(DBI)
library(tidyverse)
library(dplyr)  
library(pheatmap)
library(ggplot2)
library(reshape2)
library(lme4)     # for linear mixed models
library(lmerTest) # for p-values in lmer
library(emmeans)  # for posthoc comparisons
library(ggflowchart)
library(car)
library(sjPlot)
library(webshot)
library(stargazer)
library(factoextra)
library(FactoMineR)
library(vegan)

```

```{r include=FALSE}

db <- dbConnect(MariaDB(), dbname="CLONEID", 
                host="cloneid.cswgogbb5ufg.us-east-1.rds.amazonaws.com", 
                user="danielolofin",
                  password="Karyotyping_Analysis")

```

```{r include=FALSE}
## this is the database containing all the data. I use R to work with it but if you have experience with SQL database interactions
## you can access with any other software you like
dbListTables(db)

```

```{r include=FALSE}
dbListFields(db,"Perspective") ##karyotype data is in here.
```

```{r include=FALSE}
dbListFields(db,"Passaging") ## a lot of the metadata is here
#> dbListFields(db,"Passaging")
```

```{r include=FALSE}
dbListFields(db,"Media") ## Media info, the id column here matches the media column inside passaging.#
```

```{r include=FALSE}
## this is how to pull the karyotype data from the perspective table. For this you'll need to know the origin.
## Perhaps vural can provide a list of all the origins that belong in the experiment. "Origin" maps to "id" field in the 
## passaging table.

q <- paste0("SELECT * FROM Perspective WHERE origin='SNU-668_G1_A10_seed' AND whichPerspective='GenomePerspective'")
rs <- dbSendQuery(db, q)
```

```{r include=FALSE}
# Single query to pull all relevant profiles
q <- paste0("SELECT * FROM Perspective WHERE origin='SNU-668_G1_A10_seed' AND whichPerspective='GenomePerspective'")
rs <- dbSendQuery(db, q)
res <- dbFetch(rs)
dbClearResult(rs)

# Convert each BLOB to a double vector
kvecs <- lapply(res$profile, function(p) {
  raw_vec <- p
  readBin(raw_vec, what = "double", n = length(raw_vec) / 8, endian = "big")
})

# Stack into a matrix
kmat <- do.call(rbind, kvecs)
rownames(kmat) <- res$cloneID
```

```{r include=FALSE}
# Define origins and build query
origins <- c(
  "SNU-668_R1_A55_seedT1", "SNU-668_R3_A55_seedT1", "SNU-668_K1_A34_seedT1",
  "SNU-668_K2_A34_seedT1", "SNU-668_K3_A34_seedT1", "HGC-27_RevG_A9-F_seed",
  "HGC-27_RevG2_A7-F_seed", "SNU-668_G1_A4_seed", "SNU-668_G2_A4_seed",
  "HGC-27_RevC_A45_seed", "SNU-668_RevG_A3_seed", "SNU-668_P2_A18k_seed",
  "SNU-668_P0_A11K_seed", "SNU-668_r2_A55_seedT1", "SNU-668_G1_A10_seed",
  "SNU-668_G2_A10_seed", "SNU-668_RevG_A10_seed", "HGC-27_G1_A9-F_seed",
  "HGC-27_G2_A9_harvesT155", "HGC-27_G2_A9-F_seed", "HGC-27_G1_A9_harvesT155",
  "HGC-27_RevG_A9_harvesT149", "HGC-27_RevG2_A8_seed", "MDA-MB-231_P_A2_harvest",
  "HGC-27_C_A45_seed", "SNU-668_RevC_A27_seed", "SNU-668_C_A24_seed",
  "SNU-668_C_A4_seed", "SNU-668_RevC_A3_seed", "MDA-MB-231_RPMI_A12_VT2_seed",
  "SNU-668_P1_A18K_seed", "SNU-668_O1_A3_exp_seedT4", "SNU-668_O2_A3_exp_seedT4",
  "HGC-27_P1_A18K_harvest", "HGC-27_P2_A18K_harvest", "SUM159_NLS_4N_O1_A7K_harvest",
  "SUM159_NLS_2N_O1_A7K_harvest", "SUM-159_NLS_4N_A5M_K_harvest",
  "SUM159_NLS_4N_O2_A7K_harvest", "SUM159_NLS_2N_O2_A7K_harvest",
  "SUM-159_NLS_2N_A7M_K_harvest"
)


origin_clause <- paste0("'", origins, "'", collapse = ", ")
 q <- paste0("SELECT * FROM Perspective WHERE origin IN (", origin_clause, ") AND whichPerspective='GenomePerspective'")



# Fetch data from the database
if (!dbIsValid(db)) {
  db <- dbConnect(...)  
}
rs <- dbSendQuery(db, q)
res <- dbFetch(rs)
dbClearResult(rs)

#  Decode the BLOBs into double vectors
kvecs <- lapply(res$profile, function(p) {
  raw_vec <- p
  readBin(raw_vec, what = "double", n = length(raw_vec) / 8, endian = "big")
})

```

```{r include=FALSE}
# Building the matrix
kmat <- do.call(rbind, kvecs)
rownames(kmat) <- res$cloneID



res$treatment <- ifelse(grepl("_G1", res$origin), "G1",
              ifelse(grepl("_G2", res$origin), "G2",
              ifelse(grepl("_C", res$origin), "Control",
              ifelse(grepl("_P1", res$origin), "P1",
              ifelse(grepl("_P2", res$origin), "P2",
              ifelse(grepl("_RevC", res$origin), "RevC",
              ifelse(grepl("_RevG", res$origin), "RevG",
              ifelse(grepl("_RevG2", res$origin), "RevG2",
              ifelse(grepl("_O1", res$origin), "O1",
              ifelse(grepl("_O2", res$origin), "O2",
              ifelse(grepl("_P0", res$origin), "P0",
              ifelse(grepl("_R1", res$origin), "R1",
              ifelse(grepl("_R2", res$origin), "R2",
              ifelse(grepl("_R3", res$origin), "R3",
              ifelse(grepl("_K1", res$origin), "K1",
              ifelse(grepl("_K2", res$origin), "K2",
              ifelse(grepl("_K3", res$origin), "K3",
              ifelse(grepl("_P", res$origin), "P",
              ifelse(grepl("_r2", res$origin), "R2",
              ifelse(grepl("_K", res$origin), "K",
              ifelse(grepl("_RPMI", res$origin), "RPMI",
                     NA)))))))))))))))))))))

# Combining matrix and group metadata
karyo_df <- as.data.frame(kmat)
karyo_df$treatment <- res$treatment
karyo_df$cloneID <- res$cloneID
karyo_df$line<- res$sampleSource
karyo_df$replicate_id<- res$origin


```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyr)
library(ggplot2)

# Convert to long format for ggplot2
karyo_long <- karyo_df %>%
  pivot_longer(cols = starts_with("V"),
               names_to = "chromosome",
               values_to = "count")

# Boxplot for all chromosomes, faceted
ggplot(karyo_long, aes(x = treatment, y = count, fill = treatment)) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~chromosome, scales = "free_y", ncol = 6) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Chromosome Counts by Treatment",
       y = "Chromosome Count")

```

```{r echo=FALSE}
# Compute total chromosome per cell
karyo_df$total_chr <- rowSums(karyo_df[, paste0("V", 1:22)])

# Histogram of total chromosome count
ggplot(karyo_df, aes(x = total_chr, fill = treatment)) +
  geom_histogram(binwidth = 1, color = "black", position = "identity", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Total Chromosome Count per Cell",
       x = "Total Chromosomes",
       y = "Number of Cells")

```

```{r}
# Mean per chromosome per treatment
mean_chr <- karyo_long %>%
  group_by(treatment, chromosome) %>%
  summarise(mean_count = mean(count, na.rm = TRUE), .groups = "drop")

ggplot(mean_chr, aes(x = chromosome, y = mean_count, fill = treatment)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Mean Chromosome Count per Treatment",
       y = "Mean Count")

```

```{r}
library(pheatmap)

# Make a cell-by-chromosome matrix
karyo_mat <- as.matrix(karyo_df[, paste0("V", 1:22)])
rownames(karyo_mat) <- karyo_df$cloneID

# Simple heatmap
pheatmap(karyo_mat[1:50, ], cluster_rows = TRUE, cluster_cols = TRUE, 
         main = "Chromosome Profiles (Subset of Cells)",
         show_rownames = FALSE)

```

```{r include=FALSE}
# Load necessary libraries
library(vegan)
library(dplyr)
library(ggplot2)
  
# Step 1: Create the matrix of chromosome features
karyo_mat <- as.matrix(karyo_df[, paste0("V", 1:22)])

# Step 2: Collapse cell-level karyotypes into flask-level centroids (mean chromosome profile)
flask_centroids <- rowsum(karyo_mat, group = karyo_df$replicate_id) / ø
                   as.vector(table(karyo_df$replicate_id))

# Step 3: Create unique metadata at the flask level
meta_flask <- karyo_df %>%
  select(replicate_id, treatment, line) %>%
  distinct()

# Step 4: Compute pairwise distance matrix between flask karyotype profiles
dist_flask <- vegdist(flask_centroids, method = "manhattan")
```

```{r}
# Step 5: Run PERMANOVA — testing treatment effect after accounting for cell line differences
adonis2(dist_flask ~ line + treatment,
        data = meta_flask,
        strata = meta_flask$line,      # Keep permutations within each line
        permutations = 999)
```

**Interpretation:** In this analysis, we tested whether different treatment conditions led to statistically significant changes in karyotype composition across independently cultured flasks, while controlling for variation due to cell lines. The PERMANOVA result (F = 3.296, p = 0.638, $R^{2}$ = 0.767) indicates that although the model explains a large proportion of the variance in karyotype profiles (77%), this variation is primarily attributed to inherent differences between cell lines rather than treatment effects. The high p-value (\> 0.05) suggests there is no statistically significant treatment effect on global karyotype composition at the flask level. Thus, under the current design and data structure, the treatments applied did not induce reproducible, detectable changes in the overall chromosome profiles when adjusting for baseline variability across cell lines.

-   2: GLMM for Per-Chromosome Changes

```{r}
library(lme4)

# Binary outcome: gain defined as >2 copies
karyo_df$gain_chr08 <- ifelse(karyo_df$V8 > 2, 1, 0)

# GLMM: treatment as fixed effect, replicate as random effect
model_chr8 <- glmer(gain_chr08 ~ treatment + (1 | replicate_id), 
                    data = karyo_df, family = binomial)

summary(model_chr8)

```

```{r}
# Step 6: Test if treatment affects karyotype variability (dispersion)
disp_test <- betadisper(dist_flask, group = meta_flask$treatment, bias.adjust = TRUE)
anova(disp_test)

```

**Intepretation:** We assessed the assumption of homogeneity of multivariate dispersions using the betadisper() function. The analysis revealed no significant difference in the variability of karyotype composition across treatment groups (F(9, 22) = 0.94, p = 0.513), indicating that group dispersions are comparable. This supports the validity of our PERMANOVA results, as the observed group differences are unlikely to be driven by unequal dispersion.

```{r}
# Step 7: Visualize results using PCoA (ordination plot)
pcoa <- cmdscale(dist_flask, eig = TRUE, k = 2)
pcoa_df <- data.frame(PC1 = pcoa$points[,1],
                      PC2 = pcoa$points[,2],
                      treatment = meta_flask$treatment,
                      line = meta_flask$line)

ggplot(pcoa_df, aes(x = PC1, y = PC2, color = treatment, shape = line)) +
  geom_point(size = 4) +
  theme_minimal() +
  ggtitle("PCoA of Karyotype Profiles (Flask Centroids)")

```

```{r}
# Load libraries
library(ggplot2)
library(vegan)
library(dplyr)

# Assume these data frames are already created:
# - flask_centroids: the averaged chromosome profile for each replicate_id
# - meta_flask: metadata with columns replicate_id, line, treatment

# Merge centroids and metadata
flask_data <- cbind(meta_flask, flask_centroids)

# Scale chromosome features
scaled_centroids <- scale(flask_centroids)

#  PCA
pca_res <- prcomp(scaled_centroids)

# Add PCA scores to data frame
pca_df <- as.data.frame(pca_res$x[, 1:2])
pca_df <- cbind(meta_flask, pca_df)

# Plot PCA
ggplot(pca_df, aes(x = PC1, y = PC2, color = treatment, shape = line)) +
  geom_point(size = 4) +
  labs(title = "PCA of Karyotype Profiles (Flask Centroids)",
       x = "PC1", y = "PC2") +
  theme_minimal()

```

```{r}
# MDS using Manhattan Distance

# Compute Manhattan distance
dist_flask <- vegdist(flask_centroids, method = "manhattan")

# Classical MDS (PCoA)
mds <- cmdscale(dist_flask, k = 2, eig = TRUE)
mds_df <- data.frame(MDS1 = mds$points[,1], MDS2 = mds$points[,2])
mds_df <- cbind(meta_flask, mds_df)

# Plot MDS
ggplot(mds_df, aes(x = MDS1, y = MDS2, color = treatment, shape = line)) +
  geom_point(size = 4) +
  labs(title = "MDS of Karyotype Profiles (Manhattan Distance)",
       x = "MDS1", y = "MDS2") +
  theme_minimal()+
  theme()

```

## Findings:

Following best practices for single-cell karyotype analysis, we aggregated profiles at the replicate (flask) level and performed PERMANOVA to assess whether treatment influences global karyotype composition, adjusting for baseline differences across cell lines and restricting permutations within lines. To validate the test’s assumptions, we assessed the homogeneity of dispersion using betadisper(). The result (F = 0.94, p = 0.513) indicated no significant difference in within-group variability across treatments, confirming that PERMANOVA assumptions were met. Thus, any detected differences—or lack thereof—in karyotype composition across treatments can be confidently interpreted as genuine shifts in group centroids rather than variability artifacts.

# Next Step:

1.  Fit GLMMs on individual chromosome gains/losses

-   **Reason:** PERMANOVA looks at overall karyotype composition, but it can miss changes that occur only in specific chromosomes. So we fit Generalized Linear Mixed Models (GLMMs) to ask: **Does treatment X increase the odds of gaining chromosome 8, for example?**

```{r}
library(lme4)
library(dplyr)

# Example for Chromosome 8 gain (assuming you have a binary column)
# Replace 'gain_chr08' with the correct column name in your data
glmm_chr8 <- glmer(gain_chr08 ~ treatment + (1 | replicate_id), 
                   data = karyo_df, 
                   family = binomial)

summary(glmm_chr8)

```

**Intepretation:** The GLMM analysis of chromosome 8 gain across treatments, while adjusting for flask-level variation, shows a statistically significant reduction in gain probability under treatment K ($\rho$ = 0.048). Other treatments did not show statistically significant effects. A notable variance among replicate flasks ($\sigma^{2}$ = 3.525) supports the use of a mixed model. However, the model failed to converge fully, suggesting potential overparameterization or sparse group sizes—this warrants careful reevaluation or model simplification.

2.  Subgroup Analysis: Within a Cell Line (e.g., SNU-668 only) -**Reason:** Our original design included multiple cell lines. If those lines respond differently to treatment, a global model can hide important differences. By running PERMANOVA within each line, we ask:**Does the treatment shift karyotype composition in SNU-668 (or HGC-27) specifically?**

```{r}
for (cell_line in unique(karyo_df$line)) {
  cat("\n\n### Cell Line:", cell_line, "###\n")
  line_df <- subset(karyo_df, line == cell_line)
  line_mat <- vegdist(as.matrix(line_df[, 1:22]), method = "manhattan")
  
  print(adonis2(line_mat ~ treatment, 
                data = line_df, 
                permutations = 999))
}

```
**Interpretation:**
We performed separate PERMANOVA analyses within each cell line to assess whether treatment influenced global karyotype composition. The results revealed statistically significant differences across treatment groups in HGC-27 (R² = 0.179, F = 8.03, p = 0.001), MDAMB231 (R² = 0.197, F = 9.83, p = 0.001), and SNU-668 (R² = 0.217, F = 7.09, p = 0.001), suggesting that treatment was associated with distinct karyotypic profiles in these lines. In contrast, no significant treatment effect was observed in SUM-159 (R² = 0.033, F = 1.77, p = 0.149). These findings indicate that treatment-induced karyotypic changes are cell line–specific, warranting further investigation into line-specific responses and the chromosomal drivers of these alterations.

3.  Visualize: PCA and Distance Boxplot

```{r}
library(ggplot2)

# PCA
pca_res <- prcomp(karyo_df[, 1:22], scale. = TRUE)
pca_df <- data.frame(pca_res$x, treatment = karyo_df$treatment)

ggplot(pca_df, aes(x = PC1, y = PC2, color = treatment)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA of Karyotype Profiles")

# Distances to group centroids (for dispersion visualization)
bd <- betadisper(vegdist(as.matrix(karyo_df[, 1:22])), group = karyo_df$treatment)
boxplot(bd, main = "Dispersion of Karyotype Profiles by Treatment")

```

4.  Test Line × Treatment Interaction

-   **Reason:** We wanted to know: **Does the effect of treatment vary depending on the cell line?**

```{r}
# Assuming replicate-level aggregation
adonis2(dist_flask ~ line * treatment, 
        data = meta_flask, 
        permutations = 999, 
        strata = meta_flask$line)

```
**Interpretation:** A PERMANOVA model including the interaction between cell line and treatment was conducted to investigate whether treatment effects on karyotype composition differ by cell line. The model explained approximately 79.3% of the variance (R² = 0.793), but the interaction effect was not statistically significant (F = 1.84, p = 0.604). This suggests that, after accounting for baseline differences between cell lines, there is no strong evidence of a treatment × cell line interaction in shaping overall karyotypic profiles at the replicate (flask) level.



## Summary of my Finding based on your feedbacks and further investigation.

Our analysis explored whether cancer treatments cause statistically significant changes in single-cell karyotype profiles across multiple cell lines. The global PERMANOVA using flask-level centroids did not detect a significant treatment effect across all lines (p = 0.638), and the line × treatment interaction was also not significant (p = 0.604), suggesting no consistent treatment-driven shift in karyotype profiles when accounting for inter-line variability. However, subgroup PERMANOVA within individual cell lines revealed strong and significant treatment effects in HGC-27, MDAMB231, and SNU-668, but not in SUM-159. This pattern indicates that the effect of treatment on karyotype composition is cell line–specific and may be diluted or masked in global models due to high inter-line heterogeneity.


```{r}
library(DiagrammeR)

grViz("
digraph karyo_analysis_decision {
  
  graph [layout = dot, rankdir = TB]

  node [shape = box, style = filled, fillcolor = lightblue, fontname = Helvetica, fontsize = 10]

  # Step nodes
  A [label = 'START: Karyotype Data (22 chr per cell)\n+ Treatment + Line + Replicate']
  B [label = 'Initial Data Cleaning & Tidy Format']
  C [label = 'Exploratory Analysis\n(PCA, Boxplots, Dispersion Test)']
  D [label = 'Global PERMANOVA\nUsing Centroid of Each Replicate']
  E [label = 'Is PERMANOVA Significant?\np > 0.05']
  F1 [label = 'NO\n→ Explore by Cell Line (Subgroup PERMANOVA)']
  F2 [label = 'YES\n→ Proceed to Post-hoc/Pairwise Comparisons']
  G [label = 'Subgroup PERMANOVA by Cell Line']
  H [label = 'Any Significant Treatment Effect within Cell Lines?']
  I1 [label = 'YES\n→ Treatment Alters Karyotype in Specific Lines']
  I2 [label = 'NO\n→ Explore Individual Features']
  J [label = 'Chromosome-Level GLMM (e.g. Chr08 Gain)']
  K [label = 'Interpret Model Results\nFixed & Random Effects']
  L [label = 'REPORT:\n• Global R² and p-value\n• Subgroup Insights\n• Chromosome-Level Findings']

  # Edges
  A -> B -> C -> D -> E
  E -> F1 [label = 'p > 0.05']
  E -> F2 [label = 'p < 0.05']
  F1 -> G -> H
  H -> I1 [label = 'p < 0.05']
  H -> I2 [label = 'p > 0.05']
  I2 -> J -> K
  I1 -> L
  K -> L
  F2 -> L
}
")


```



```{r}
query <- "SELECT id, passaged_from_id1, passage, event, date, media from Passaging WHERE passaged_from_id1 != 'NA'"
mismatches <- dbGetQuery(db, query)


# --- Comparison 1: Passage Mismatch ---
# Perform a self-join on the 'id' and 'passaged_from_id1' columns to compare passage numbers
# Here, we are joining the data on the id = passaged_from_id1 to compare passages
comparisonSeed <- mismatches[mismatches$event=='seeding',] %>%
  inner_join(mismatches, by = c("passaged_from_id1" = "id"), suffix = c("_new", "_original")) %>%
  filter(passage_new != passage_original + 1)

# 'comparisonSeed' now contains rows where the passage of the new id is not equal to 
# the passage of the passaged_from_id1 + 1
# add harvest comparison where passage_new != passage_original
comparisonHarvest <- mismatches[mismatches$event=='harvest',] %>%
  inner_join(mismatches, by = c("passaged_from_id1" = "id"), suffix = c("_new", "_original")) %>%
  filter(passage_new != passage_original)

# Add a column to label this comparison
passage_comparison <- bind_rows(comparisonSeed ,comparisonHarvest)
passage_comparison$comparison_type <- "Passage Mismatch"
View(comparisonSeed)

# --- Comparison: Name Mismatch Before "_A#" ---
# Extract the parts of id and passaged_from_id1 before "_A#"
# Remove hyphens/dashes and compare the extracted parts
comparisonID <- mismatches %>%
  mutate(
    id_clean = gsub("-", "", sub("_A\\d+.*", "", id)),  # Remove hyphens, trim from "_A#" onward
    passaged_from_id1_clean = gsub("-", "", sub("_A\\d+.*", "", passaged_from_id1))
  ) %>%
  filter(id_clean != passaged_from_id1_clean)  # Keep only mismatches

# Add a label column
comparisonID$comparison_type <- "Name Mismatch"

# View result
View(comparisonID)
# Add a column to label this comparison
comparisonID$comparison_type <- "Name Mismatch"


extract_passage_number <- function(x) {
  # Use regular expression to extract the number following '_A'
  match <- str_match(x, "_A(\\d+)")
  
  # If there is a match, return the number as integer, otherwise return NA
  if (!is.na(match[2])) {
    return(as.numeric(match[2]))
  } else {
    return(NA)
  }
}

# Apply the A# function to both id and passaged_from_id1 columns
mismatches$passage_num_id <- sapply(mismatches$id, extract_passage_number)
mismatches$passage_num_from <- sapply(mismatches$passaged_from_id1, extract_passage_number)

# Filter out rows where the passage numbers are equal (i.e., no mismatch)
harvest_passageID <- mismatches[mismatches$event=='harvest',] %>%
  filter(!is.na(passage_num_id) & !is.na(passage_num_from) & passage_num_id != passage_num_from)
seed_passageID <- mismatches[mismatches$event=='seeding',] %>%
  filter(!is.na(passage_num_id) & !is.na(passage_num_from) & passage_num_id != 1+passage_num_from)
passageID_comparison <- bind_rows(seed_passageID, harvest_passageID) 

# Add a column to label this comparison
passageID_comparison$comparison_type <- "Passage ID Mismatch"


# --- Comparison 3: Media Mismatch ---
# Perform self-join to align the media of `id` with `passaged_from_id1`
media_comparison <- mismatches %>%
  inner_join(mismatches, by = c("passaged_from_id1" = "id"), suffix = c("_id", "_from")) %>%
  filter(media_id != media_from)

# Add a column to label this comparison as "Media Mismatch"
media_comparison$comparison_type <- "Media Mismatch"
write.csv(comparisonID, "~/Documents/R_codes/name_comparison.csv", row.names = FALSE)


# --- Combine Results ---
# Select only relevant columns for all comparisons
# passage_comparison <- passage_comparison %>% select(id, passaged_from_id1, passage, comparison_type)
# comparisonID <- name_comparison %>% select(id, passaged_from_id1, comparison_type)
# media_comparison <- media_comparison %>% select(id, passaged_from_id1, media, comparison_type)

# Combine all comparisons into one data frame
combined_comparisons <- bind_rows(passage_comparison, comparisonID, media_comparison, passageID_comparison)
# Remove duplicate entries based on the "id" column
combined_comparisons_unique <- combined_comparisons %>%
  distinct(id, .keep_all = TRUE)

# Write the combined unique results to a CSV file
write.csv(combined_comparisons_unique, "~/Documents/R_codes/combined_comparisons_unique1.csv", row.names = FALSE)


# Filter out rows where the passage numbers are equal (i.e., no mismatch)
mismatch_final <- mismatches[!is.na(mismatches$passage_num_id) & 
                               !is.na(mismatches$passage_num_from) & 
                               mismatches$passage_num_id != mismatches$passage_num_from, ]
mismatch_final$suggested_corrections <- ""
# Print the mismatched rows
View(mismatch_final)

#write.csv(mismatch_harvest, "mismatch_data.csv", row.names = FALSE)
write.csv(mismatch_final, "~/Documents/R_codes/mismatch_data.csv", row.names = FALSE)

dbDisconnect(db)
```



######################## g


```{r eval=FALSE, include=FALSE}
q <- paste0("SELECT * FROM Passaging")
rs <- dbSendQuery(db, q)
res <- dbFetch(rs)
dbClearResult(rs)


# Joining the passaging data 
joined_data <- left_join(karyo_df, res_passage, by = "replicate_id")


# Creating a new data set for SNU_668
joined_data %>% 
  filter(cellLine == "SNU-668" & treatment == "Control") -> SNU_668

round(SNU_668$date[33] %--% SNU_668$date[1]/months(1))
```






```{r}
# Load required libraries
library(tidyverse)

# Step 1: Prepare baseline (control) reference profile
# Assuming karyo_df is your real dataset with chromosomes V1 to V22
# And it contains "treatment" column
control_profile <- karyo_df %>%
  filter(line == "SNU-668",treatment == "Control") %>%
  select(starts_with("V")) %>%
  summarise(across(everything(), mean, na.rm = TRUE)) %>%
  unlist()

# Step 2: Define simulation function
simulate_karyotypes <- function(n_cells = 1000, n_chrom = 22, time_steps = 10, p_gain = 0.03, p_loss = 0.03) {
  karyo_list <- vector("list", n_cells)
  for (i in 1:n_cells) {
    # Start with random cancer-like karyotype (non-diploid) between 0–5
    karyo <- sample(0:5, n_chrom, replace = TRUE)
    for (t in 1:time_steps) {
      change <- sample(c(1, -1, 0), n_chrom, replace = TRUE,
                       prob = c(p_gain, p_loss, 1 - p_gain - p_loss))
      karyo <- pmax(0, karyo + change)
    }
    karyo_list[[i]] <- karyo
  }
  sim_mat <- do.call(rbind, karyo_list)
  colnames(sim_mat) <- paste0("V", 1:n_chrom)
  return(sim_mat)
}

# Step 3: Define objective function to minimize
objective_fn <- function(params) {
  p_gain <- params[1]
  p_loss <- params[2]
  sim <- simulate_karyotypes(p_gain = p_gain, p_loss = p_loss)
  sim_profile <- colMeans(sim)
  dist <- sum(abs(sim_profile - control_profile)) # Manhattan distance
  return(dist)
}

# Step 4: Optimize
set.seed(42)
opt <- optim(par = c(0.03, 0.03), fn = objective_fn, method = "L-BFGS-B",
             lower = c(0.01, 0.01), upper = c(0.2, 0.2))

opt_params <- opt$par
cat("Optimal P_gain:", opt_params[1], "\n")
cat("Optimal P_loss:", opt_params[2], "\n")

# Step 5: Final simulation with optimized params
optimized_sim <- simulate_karyotypes(n_cells = 1000,
                                     p_gain = opt_params[1],
                                     p_loss = opt_params[2])

# Step 6: Show simulated result
sim_df <- as.data.frame(optimized_sim)
sim_df$cell <- paste0("Cell_", 1:nrow(sim_df))

# Show first few rows
head(sim_df)

```


```{r}
library(tidyverse)
library(ggplot2)
library(pheatmap)
library(dplyr)
library(tidyr)
library(scales)
```

## Theoretical Framework

In this study, we aim to model **karyotype evolution in cancer cells (SNU-668 cell line)** under untreated (control) conditions over a time span of four months. The motivation stems from distinguishing **stochastic drift** from **directed karyotypic evolution** by quantifying changes in chromosomal copy number across time.

### Baseline and Target Definitions
- **Baseline**: The replicate `SNU-668_C_A4_seed`, representing the early control state (Month 0).
- **Target**: The replicate `SNU-668_C_A24_seed`, representing the evolved control karyotype after 4 months (Month 4).

We simulate chromosomal copy number evolution over multiple time steps to match the late control profile. The simulation framework is built on a **discrete-time Markov chain model**, assuming each chromosome in each cell can independently:

- Gain a copy (→ +1) with probability \( p_\text{gain} \)
- Lose a copy (→ −1) with probability \( p_\text{loss} \)
- Remain unchanged with probability \( 1 - p_\text{gain} - p_\text{loss} \)

Copy numbers are initialized based on the actual observed `SNU-668_C_A4_seed` data. Over a range of simulation steps, we generate predicted karyotype profiles and compute distances to the `SNU-668_C_A24_seed` profile. The **optimal number of time steps** is chosen as the one that minimizes this distance.

### Objective
To identify the number of time steps (or simulation length) that best mimics the empirical karyotype drift between early and late control replicates.

---

## Load Input Reference (Early and Late Control)
```{r load-control-karyo}
# Mock-up: Replace this with actual data
# karyo_df should be a data frame with columns V1:V22, treatment, line, replicate_id

# Filter for SNU-668 control replicates
karyo_control <- karyo_df %>% filter(line == "SNU-668", treatment == "Control")
early_control <- karyo_control %>% filter(replicate_id == "SNU-668_C_A4_seed")
late_control <- karyo_control %>% filter(replicate_id == "SNU-668_C_A24_seed")

# Aggregate mean chromosome values per replicate
early_avg <- early_control %>% select(starts_with("V")) %>% summarise_all(mean)
late_avg  <- late_control %>% select(starts_with("V")) %>% summarise_all(mean)
```

## Define Stochastic Simulation Function
```{r simulation-function}
simulate_karyotype <- function(initial_vec, n_cells = 1000, steps = 1, p_gain = 0.03, p_loss = 0.03) {
  n_chrom <- length(initial_vec)
  results <- matrix(rep(initial_vec, each = n_cells), nrow = n_cells)
  for (t in 1:steps) {
    delta <- replicate(n_chrom, sample(c(1, -1, 0), n_cells, replace = TRUE, prob = c(p_gain, p_loss, 1 - p_gain - p_loss)))
    results <- pmax(0, results + delta)
  }
  colnames(results) <- names(initial_vec)
  as.data.frame(results)
}
```

## Run Simulations Over Time Steps
```{r run-simulations}
time_steps <- 1:30

# Store distances at each time point
distance_profile <- data.frame(time = time_steps, distance = NA)

for (t in time_steps) {
  sim <- simulate_karyotype(as.numeric(early_avg), steps = t)
  sim_avg <- colMeans(sim)
  dist <- sum(abs(sim_avg - as.numeric(late_avg)))
  distance_profile$distance[t] <- dist
}

# Plot optimization
ggplot(distance_profile, aes(x = time, y = distance)) +
  geom_line(size = 1.2) +
  geom_point() +
  labs(title = "Optimization: Simulated vs Observed Late Control",
       x = "Simulation Time Steps",
       y = "Manhattan Distance to Late Control") +
  theme_minimal()
```

## Visualize Final Simulation
```{r final-simulation-heatmap}
# Simulate for optimal step
best_time <- distance_profile$time[which.min(distance_profile$distance)]
sim_final <- simulate_karyotype(as.numeric(early_avg), steps = best_time)

# Heatmap (random subset of cells)
heat_mat <- as.matrix(sim_final[1:50,])
pheatmap(heat_mat, cluster_rows = TRUE, cluster_cols = TRUE,
         main = paste("Simulated Karyotype (Control - Time Step", best_time, ")"),
         show_rownames = FALSE)

```






## Theoretical Framework

We model the chromosomal evolution of cancer cells using a **discrete-time stochastic process**, specifically a **Markov chain** where each chromosome in a given cell may independently:

- Gain a copy (+1),
- Lose a copy (−1), or
- Remain unchanged (0)

at each time step.

Let:
- $C_{i,t}$ be the copy number of chromosome $i$ at time $t$,
- $P_g$ be the probability of gaining a chromosome,
- $P_l$ be the probability of losing a chromosome.

### Transition Rule
$$
C_{i,t+1} = \begin{cases}
  C_{i,t} + 1 & \text{with probability } P_g \\
  C_{i,t} - 1 & \text{with probability } P_l, \text{ if } C_{i,t} > 0 \\
  C_{i,t}     & \text{with probability } 1 - P_g - P_l
\end{cases}
$$

We define the initial state $C_{i,0}$ using the early control replicate (`SNU-668_C_A4_seed`) and optimize $P_g$ and $P_l$ such that the simulated average profile after 4 months best approximates the late control replicate (`SNU-668_C_A24_seed`).

## Load and Prepare Data
```{r}
# Assume karyo_df is already loaded and contains columns V1 to V22, treatment, replicate_id
baseline <- karyo_df %>%
  filter(line == "SNU-668", replicate_id == "SNU-668_C_A4_seed") %>%
  select(starts_with("V")) %>%
  summarise(across(everything(), mean)) %>%
  unlist()

late_control <- karyo_df %>%
  filter(line == "SNU-668", replicate_id == "SNU-668_C_A24_seed") %>%
  select(starts_with("V")) %>%
  summarise(across(everything(), mean)) %>%
  unlist()
```

## Simulation Function
```{r}
simulate_karyotypes <- function(init_profile, n_cells = 1000, time_steps = 10, p_gain = 0.03, p_loss = 0.03) {
  n_chrom <- length(init_profile)
  sim_list <- vector("list", n_cells)

  for (i in 1:n_cells) {
    karyo <- round(init_profile)  # Start with early control
    for (t in 1:time_steps) {
      change <- sample(c(+1, -1, 0), size = n_chrom, replace = TRUE,
                       prob = c(p_gain, p_loss, 1 - p_gain - p_loss))
      karyo <- pmax(0, karyo + change)
    }
    sim_list[[i]] <- karyo
  }
  do.call(rbind, sim_list)
}
```

## Objective Function (Distance to Late Control)
```{r}
objective_fn <- function(params) {
  p_gain <- params[1]
  p_loss <- params[2]
  sim <- simulate_karyotypes(init_profile = baseline, p_gain = p_gain, p_loss = p_loss)
  sim_profile <- colMeans(sim)
  dist <- sum(abs(sim_profile - late_control)) # Manhattan distance
  return(dist)
}
```

## Optimization
```{r}
set.seed(42)
opt <- optim(par = c(0.03, 0.03), fn = objective_fn, method = "L-BFGS-B",
             lower = c(0.01, 0.01), upper = c(0.2, 0.2))
opt_params <- opt$par
cat("Optimal P_gain:", opt_params[1], "\n")
cat("Optimal P_loss:", opt_params[2], "\n")
```

## Final Simulation
```{r}
final_sim <- simulate_karyotypes(init_profile = baseline,
                                 p_gain = opt_params[1],
                                 p_loss = opt_params[2])
```

## Heatmap Visualization
```{r}
heat_mat <- final_sim %>%
  as.data.frame() %>%
  mutate(cell = paste0("Cell_", 1:nrow(.))) %>%
  column_to_rownames("cell") %>%
  select(paste0("V", 1:22))

pheatmap(heat_mat[1:50, ],
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         main = "Simulated Chromosome Profiles (First 50 Cells)",
         show_rownames = FALSE)
```

## Distance Between Simulation and Late Control
```{r}
final_profile <- colMeans(final_sim)
distance_to_late <- sum(abs(final_profile - late_control))
cat("Final Manhattan distance to late control:", distance_to_late)
```




### gggg

```{r}
library(tidyverse)

# Step 1: Extract early and late control cell-level karyotypes
early_control_mat <- karyo_df %>%
  filter(replicate_id == "SNU-668_C_A4_seed") %>%
  select(starts_with("V")) %>%
  as.matrix()

late_control_profile <- karyo_df %>%
  filter(replicate_id == "SNU-668_C_A24_seed") %>%
  select(starts_with("V")) %>%
  summarise(across(everything(), mean, na.rm = TRUE)) %>%
  unlist()

# Step 2: Define simulation function (randomly sample start profiles from early pool)
simulate_karyotypes <- function(start_pool, p_gain, p_loss, n_cells = 1000, time_steps = 10) {
  n_chrom <- ncol(start_pool)
  karyo_list <- vector("list", n_cells)
  for (i in 1:n_cells) {
    karyo <- start_pool[sample(1:nrow(start_pool), 1, replace = TRUE), ]
    for (t in 1:time_steps) {
      change <- sample(c(1, -1, 0), size = n_chrom, replace = TRUE,
                       prob = c(p_gain, p_loss, 1 - p_gain - p_loss))
      karyo <- pmax(0, karyo + change)
    }
    karyo_list[[i]] <- karyo
  }
  do.call(rbind, karyo_list)
}

# Step 3: Define parameter grid (grid search)
p_gain_seq <- seq(0.01, 0.2, by = 0.01)
p_loss_seq <- seq(0.01, 0.2, by = 0.01)
param_grid <- expand.grid(p_gain = p_gain_seq, p_loss = p_loss_seq)

# Step 4: Evaluate all parameter combinations (minimize distance to late control)
results <- param_grid %>%
  rowwise() %>%
  mutate(
    dist = {
      sim <- simulate_karyotypes(
        start_pool = early_control_mat,
        p_gain = p_gain,
        p_loss = p_loss,
        n_cells = 1000,
        time_steps = 10  # Simulating over 4 months
      )
      sim_profile <- colMeans(sim)
      sum(abs(sim_profile - late_control_profile))  # Manhattan distance
    }
  )

# Step 5: Find optimal parameters
best_params <- results %>% arrange(dist) %>% slice(1)
print(best_params)

# Step 6: Final simulation using optimized probabilities
final_sim <- simulate_karyotypes(
  start_pool = early_control_mat,
  p_gain = best_params$p_gain,
  p_loss = best_params$p_loss,
  n_cells = 1000,
  time_steps = 10
)

# Step 7: Visualize distribution (Corrected pivot_longer)
sim_df <- as.data.frame(final_sim)
sim_df$cell <- paste0("Cell_", 1:nrow(sim_df))

sim_long <- sim_df %>%
  select(-cell) %>%
  pivot_longer(cols = everything(), names_to = "chrom", values_to = "copy")

ggplot(sim_long, aes(x = copy)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  facet_wrap(~chrom, scales = "free_y") +
  theme_minimal() +
  labs(title = "Simulated Chromosome Copy Number After 4 Months (Control - SNU-668)",
       x = "Copy Number", y = "Frequency")
```

```{r}
heat_mat <- sim_df %>%
  mutate(cell = paste0("Cell_", 1:nrow(.))) %>%
  column_to_rownames("cell") %>%
  select(paste0("V", 1:22))

pheatmap(heat_mat[1:50, ],
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         main = "Simulated Chromosome Profiles (First 50 Cells)",
         show_rownames = FALSE)
```

