---
title: "Decoding ML"
author: "Olofin Daniel"
date: "2025-07-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Liberies
```{r message=FALSE, warning=FALSE}
library(dplyr)
library(tidyverse)
library(stringr)
library(naniar)
library(gridExtra)
library(grid)
library(mice)
library(dlookr)
```


```{r message=FALSE, warning=FALSE}
data <- read.csv("projectdata.csv")
names(data)
# Select the relevant columns
selected_data <- data[, c("Mutation.Count", 
                          "Somatic.Status", 
                          "Gene.Panel", 
                          "Fraction.Genome.Altered", 
                          "MSI.Type", 
                          "Sex", 
                          "Race", 
                          "Ethnicity", 
                          "Sample.Type", 
                          "Overall.Survival.Status",
                          "Overall.Survival..Months.")]
```
 **I read the original dataset and create a new data set with only needed variables before cleaned the data!**

## Checking data structure
```{r echo=FALSE, message=FALSE, warning=FALSE}
str(selected_data)
```

## Checking unique objject in each variable

```{r message=FALSE, warning=FALSE}
# Function to show unique values for character columns
show_unique_characters <- function(data) {
  # Loop through each column in the dataset
  for (col_name in colnames(data)) {
    # Check if the column is of type character or factor
    if (is.character(data[[col_name]]) | is.factor(data[[col_name]])) {
      cat("\nUnique values in column:", col_name, "\n")
      print(unique(data[[col_name]]))  # Show unique values
    }
  }
}

# Call the function with your dataset
show_unique_characters(selected_data)

```

## Dealing with ""
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Function to replace empty strings with NA
replace_empty_with_na <- function(data) {
  # Loop through each column in the dataset
  data[] <- lapply(data, function(x) {
    # Replace empty strings with NA
    if (is.character(x) | is.factor(x)) {
      x[x == ""] <- NA
    }
    return(x)
  })
  return(data)
}

# Call the function with your dataset
selected_data <- replace_empty_with_na(selected_data)

```


## Correcting the unique object 
```{r message=FALSE, warning=FALSE}
selected_data<-selected_data %>% 
  mutate(Sex = str_to_sentence(Sex),
         Race = str_to_sentence(Race))

selected_data %>% 
  separate(Overall.Survival.Status, into = c("Overall.Survival.Status_num","Overall.Survival.Status")) %>% 
  mutate(Overall.Survival.Status = str_to_sentence(Overall.Survival.Status),
         Gene.Panel = str_to_sentence(Gene.Panel)) -> selected_data
```


## Checking the Missing variables

```{r message=FALSE, warning=FALSE}
gg_miss_var(selected_data)

```

## Distribution of Continous variables
### Histograms or Density Plots for Continuous Variables (e.g., Mutation Count, Fraction Genome Altered)


```{r echo=FALSE, fig.height=8, fig.width=16, message=FALSE, warning=FALSE, out.width="100%"}

# Mutation Count
p1<-ggplot(selected_data, aes(x = Mutation.Count)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Mutation Count", x = "Mutation Count", y = "Frequency")+
  theme_minimal()

# Fraction Genome Altered
p2<-ggplot(selected_data, aes(x = Fraction.Genome.Altered)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Fraction Genome Altered", x = "Fraction Genome Altered", y = "Frequency")+
  theme_minimal()

# Fraction Genome Altered
p3<-ggplot(selected_data, aes(x = Overall.Survival..Months.)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Fraction Genome Altered", x = "Fraction Genome Altered", y = "Frequency")+
  theme_minimal()
# Combine the two heatmaps
grid.arrange(p1, p2, p3, ncol = 2)

```

## Boxplots for Genetic and Demographic Factors
### Boxplot of Mutation Count by Sample Type or Sex
```{r echo=FALSE, fig.height=8, fig.width=16, message=FALSE, warning=FALSE, out.width="100%"}
p3<-ggplot(selected_data, aes(x = Sample.Type, y = Mutation.Count, fill = Sample.Type)) +
  geom_boxplot(show.legend = F) +
  labs(title = "Mutation Count by Sample Type", x = "Sample Type", y = "Mutation Count")

p4<-ggplot(selected_data, aes(x = Sex, y = Mutation.Count, fill = Sex)) +
  geom_boxplot(show.legend = F) +
  labs(title = "Mutation Count by Sex", x = "Sex", y = "Mutation Count")

# Combine the two heatmaps
grid.arrange(p3, p4, ncol = 2)

```

### Boxplot of Fraction Genome Altered by Race or Ethnicity
```{r echo=FALSE, fig.height=8, fig.width=16, message=FALSE, warning=FALSE, out.width="100%"}
p5<-ggplot(selected_data, aes(x = Race, y = Fraction.Genome.Altered, fill = Race)) +
  geom_boxplot(show.legend = F) +
  labs(title = "Fraction Genome Altered by Race", x = "Race", y = "Fraction Genome Altered")

p6<-ggplot(selected_data, aes(x = Ethnicity, y = Fraction.Genome.Altered, fill = Ethnicity)) +
  geom_boxplot(show.legend = F) +
  labs(title = "Fraction Genome Altered by Ethnicity", x = "Ethnicity", y = "Fraction Genome Altered")

# Combine the two heatmaps
grid.arrange(p5, p6, ncol = 2)
```

### Bar Plot for Sex, Race, and Ethnicity
```{r echo=FALSE, fig.height=8, fig.width=16, message=FALSE, warning=FALSE, out.width="100%"}
p7<-ggplot(selected_data, aes(x = Sex, fill = Sex)) +
  geom_bar(show.legend = F) +
  labs(title = "Distribution of Sex", x = "Sex", y = "Count")

p8<- ggplot(selected_data, aes(x = Race, fill = Race)) +
  geom_bar(show.legend = F) +
  labs(title = "Distribution of Race", x = "Race", y = "Count")

p9<-ggplot(selected_data, aes(x = Ethnicity, fill = Ethnicity)) +
  geom_bar(show.legend = F) +
  labs(title = "Distribution of Ethnicity", x = "Ethnicity", y = "Count")

# Combine the two heatmaps
grid.arrange(p7, p8, p9, ncol = 2)
```

## Handling Missing and Unknown

```{r echo=FALSE, fig.height=8, fig.width=16, message=FALSE, warning=FALSE, out.width="100%"}
p1<- dlookr::imputate_na(selected_data,Fraction.Genome.Altered, method = "mean") %>% plot()
p2<-dlookr::imputate_na(selected_data,Fraction.Genome.Altered, method = "median") %>% plot()
p3<-dlookr::imputate_na(selected_data,Fraction.Genome.Altered, method = "mode") %>% plot()
p4<-dlookr::imputate_na(selected_data,Fraction.Genome.Altered, method = "mice") %>% plot()


# Combine the two heatmaps
grid.arrange(p1, p2, p3, p4, ncol = 2)
```
**Look at the graph above it shows that mice imputation matches well than another methods**


```{r echo=FALSE, fig.height=8, fig.width=16, message=FALSE, warning=FALSE, out.width="100%"}
p1<- dlookr::imputate_na(selected_data,Mutation.Count, method = "mean") %>% plot()
p2<-dlookr::imputate_na(selected_data,Mutation.Count, method = "median") %>% plot()
p3<-dlookr::imputate_na(selected_data,Mutation.Count, method = "mode") %>% plot()
p4<-dlookr::imputate_na(selected_data,Mutation.Count, method = "mice") %>% plot()


# Combine the two heatmaps
grid.arrange(p1, p2, p3, p4, ncol = 2)
```


```{r echo=FALSE, fig.height=8, fig.width=16, message=FALSE, warning=FALSE, out.width="100%"}
p1<- dlookr::imputate_na(selected_data,Overall.Survival..Months., method = "mean") %>% plot()
p2<-dlookr::imputate_na(selected_data,Overall.Survival..Months., method = "median") %>% plot()
p3<-dlookr::imputate_na(selected_data,Overall.Survival..Months., method = "mode") %>% plot()
p4<-dlookr::imputate_na(selected_data,Overall.Survival..Months., method = "mice") %>% plot()


# Combine the two heatmaps
grid.arrange(p1, p2, p3, p4, ncol = 2)

selected_data<-selected_data %>% 
  mutate(Fraction.Genome.Altered = imputate_na(selected_data,Fraction.Genome.Altered, method = "mice"),
        Mutation.Count = imputate_na(selected_data,Mutation.Count, method = "mice"),
        Overall.Survival..Months. =imputate_na(selected_data,Overall.Survival..Months., method = "mice"))
```

### Function for imputing na for character type

```{r message=FALSE, warning=FALSE}
# Function to impute missing values with the mode for character columns
impute_na_with_mode <- function(data) {
  # Loop through each column in the dataset
  data[] <- lapply(data, function(x) {
    # Only apply to character or factor columns
    if (is.character(x) | is.factor(x)) {
      # Calculate the mode (most frequent value)
      mode_value <- names(sort(table(x), decreasing = TRUE))[1]
      # Replace NAs with the mode value
      x[is.na(x)] <- mode_value
    }
    return(x)
  })
  return(data)
}

# Call the function with your dataset
selected_data <- impute_na_with_mode(selected_data)
```
**I wrote a function that impute all missing values in character type data with mode**


### Confirmation Missing variables

```{r message=FALSE, warning=FALSE}
gg_miss_var(selected_data)

```



### Droping all values with unknow

```{r echo=FALSE, message=FALSE, warning=FALSE}
selected_data %>% 
  group_by(Sex) %>% 
  summarise(n = n())

selected_data %>% 
  group_by(Race) %>% 
  summarise(n = n())

selected_data %>% 
  group_by(Ethnicity) %>% 
  summarise(n = n())

selected_data %>% 
  group_by(Sample.Type) %>% 
  summarise(n = n())

```

```{r}
selected_data %>% 
  filter(Sex == "Unknown" | Race == "Unknown" | Ethnicity %in% c("Unknown whether Spanish or not","Unknown") | Sample.Type == "Unknown") %>% dim()

selected_data %>% 
  filter(Sex != "Unknown" & Race != "Unknown" & 
         !Ethnicity %in% c("Unknown whether Spanish or not", "Unknown") & 
         Sample.Type != "Unknown") ->selected_data

```
**188 observation was drop because they have unkown**


# New Analysis After Data cleaning...

## Distribution of Continous variables
### Histograms or Density Plots for Continuous Variables (e.g., Mutation Count, Fraction Genome Altered)


```{r echo=FALSE, fig.height=8, fig.width=16, message=FALSE, warning=FALSE, out.width="100%"}

# Mutation Count
p1<-ggplot(selected_data, aes(x = Mutation.Count)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Mutation Count", x = "Mutation Count", y = "Frequency")+
  theme_minimal()

# Fraction Genome Altered
p2<-ggplot(selected_data, aes(x = Fraction.Genome.Altered)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Fraction Genome Altered", x = "Fraction Genome Altered", y = "Frequency")+
  theme_minimal()

# Fraction Genome Altered
p3<-ggplot(selected_data, aes(x = Overall.Survival..Months.)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Fraction Genome Altered", x = "Fraction Genome Altered", y = "Frequency")+
  theme_minimal()
# Combine the two heatmaps
grid.arrange(p1, p2, p3, ncol = 2)

```

## Boxplots for Genetic and Demographic Factors
### Boxplot of Mutation Count by Sample Type or Sex
```{r echo=FALSE, fig.height=8, fig.width=16, message=FALSE, warning=FALSE, out.width="100%"}
p3<-ggplot(selected_data, aes(x = Sample.Type, y = Mutation.Count, fill = Sample.Type)) +
  geom_boxplot(show.legend = F) +
  labs(title = "Mutation Count by Sample Type", x = "Sample Type", y = "Mutation Count")

p4<-ggplot(selected_data, aes(x = Sex, y = Mutation.Count, fill = Sex)) +
  geom_boxplot(show.legend = F) +
  labs(title = "Mutation Count by Sex", x = "Sex", y = "Mutation Count")

# Combine the two heatmaps
grid.arrange(p3, p4, ncol = 2)

```

### Boxplot of Fraction Genome Altered by Race or Ethnicity
```{r echo=FALSE, fig.height=8, fig.width=16, message=FALSE, warning=FALSE, out.width="100%"}
p5<-ggplot(selected_data, aes(x = Race, y = Fraction.Genome.Altered, fill = Race)) +
  geom_boxplot(show.legend = F) +
  labs(title = "Fraction Genome Altered by Race", x = "Race", y = "Fraction Genome Altered")

p6<-ggplot(selected_data, aes(x = Ethnicity, y = Fraction.Genome.Altered, fill = Ethnicity)) +
  geom_boxplot(show.legend = F) +
  labs(title = "Fraction Genome Altered by Ethnicity", x = "Ethnicity", y = "Fraction Genome Altered")

# Combine the two heatmaps
grid.arrange(p5, p6, ncol = 2)
```

### Bar Plot for Sex, Race, and Ethnicity
```{r echo=FALSE, fig.height=8, fig.width=16, message=FALSE, warning=FALSE, out.width="100%"}
p7<-ggplot(selected_data, aes(x = Sex, fill = Sex)) +
  geom_bar(show.legend = F) +
  labs(title = "Distribution of Sex", x = "Sex", y = "Count")

p8<- ggplot(selected_data, aes(x = Race, fill = Race)) +
  geom_bar(show.legend = F) +
  labs(title = "Distribution of Race", x = "Race", y = "Count")

p9<-ggplot(selected_data, aes(x = Ethnicity, fill = Ethnicity)) +
  geom_bar(show.legend = F) +
  labs(title = "Distribution of Ethnicity", x = "Ethnicity", y = "Count")

# Combine the two heatmaps
grid.arrange(p7, p8, p9, ncol = 2)
```

