---
title: "Deep Temporal Learning for HIV Forecasting: Modeling Epidemic Dynamics with Long Short-Term Memory Networks"
author: "Daniel Olofin"
date: "2025-10-26"
output: 
  pdf_document:
    latex_engine: xelatex
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(readxl)
library(readr)
library(keras)
library(tensorflow)

```



```{r}
HiV_data <- read_xlsx("Hivdata.xlsx")
```

```{r echo=FALSE, fig.height=7, fig.width=10, message=FALSE, warning=FALSE, out.width="100%"}
library(viridis)
library(scales)
library(tidyr)
library(tidyverse)
# Example data (replace this with your actual dataset)
df <- HiV_data

df <- df %>%
  tidyr::fill(YEAR)

# Summarize your data
df_sum <- df %>%
  group_by(YEAR) %>%
  summarise(total = sum(prevalence, na.rm = TRUE))  # replace with your actual variable

# Identify the year with the highest total
max_year <- df_sum %>%
  filter(total == max(total)) %>%
  pull(YEAR)

# Create the plot
ggplot(df_sum, aes(x = factor(YEAR), y = total,
                   fill = ifelse(YEAR == max_year, "Highest", "Other"))) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = round(total, 2)), vjust = -0.3, size = 3) +
  scale_fill_manual(values = c("Highest" = "black", "Other" = "steelblue")) +
  labs(
  title = "Annual Aggregation of HIV Prevalence for Temporal Modeling (2010–2021)",
  x = "Year",
  y = NULL
) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none",
    plot.title = element_text(face = "bold")
  ) -> p


df <- df %>%
  tidyr::fill(YEAR) %>%                                   # fill down missing YEARs
  mutate(
    Month = str_trim(Month),                       # remove leading/trailing spaces
    Month = str_to_upper(Month),                   # ensure all uppercase
    Month_num = match(Month, toupper(month.abb)),  # map JAN→1, FEB→2, ...
    date = make_date(YEAR, Month_num, 1)           # create full date
  )


# plot 
p<- ggplot(df, aes(x = Month_num, y = prevalence, group = YEAR)) +
  geom_line(color="blue") +
  geom_point(color="red") +
  facet_wrap(~ YEAR, ncol = 3, scales = "free_y") + 
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1)) +

 labs(
  title = "Temporal Dynamics of Monthly HIV Prevalence (2010–2021)",
  x = "Month",
  y = "Prevalence"
)


ggsave(
  filename = "HIV_Prevalence_Trajectory_2010_2021.png",
  plot = p,
  width = 8,        # width in inches
  height = 5,       # height in inches
  dpi = 300         # resolution (suitable for publication)
)
```

```{r echo=FALSE, fig.height=7, fig.width=10, message=FALSE, warning=FALSE, out.width="100%"}
ggplot(df, aes(x = date, y = prevalence)) +
  geom_line(color = "blue") +
geom_point(color="red") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1)) +
  labs(
  title = "Temporal Trajectory of HIV Prevalence (2010–2021)",
  x = "Date",
  y = "Prevalence"
)-> p

ggsave(
  filename = "HIV_Prevalence_Trajectory_2010_2021.png",
  plot = p,
  width = 8,        # width in inches
  height = 5,       # height in inches
  dpi = 300         # resolution (suitable for publication)
)
```


# LSTM
```{r}

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(lubridate)
  library(ggplot2)
  library(keras)
  library(tensorflow)
  library(purrr)
  library(abind)
})

# --- Reproducibility ---
set.seed(123)
tensorflow::tf$random$set_seed(123)

# --- 0) Data: ensure date is Date and sorted monthly ---
stopifnot(all(c("YEAR","Month","prevalence","Month_num","date") %in% names(df)))
df <- df %>%
  mutate(date = as.Date(date)) %>%
  arrange(date)

# Keep only modeling columns
df <- df %>% select(date, prevalence)


```


```{r}
# ================================================
# LSTM for Monthly HIV Prevalence (R / keras)
# Expects df with columns: YEAR, Month, prevalence, Month_num, date (YYYY-MM-DD)
# ================================================

suppressPackageStartupMessages({
  library(dplyr)
  library(lubridate)
  library(ggplot2)
  library(purrr)
  library(keras)
  library(tensorflow)
  library(abind)
})

# --- Reproducibility ---
set.seed(123)
tensorflow::tf$random$set_seed(123)

# --- Ensure date column and sort ---
df <- df %>%
  mutate(date = as.Date(date)) %>%
  arrange(date) %>%
  select(date, prevalence)

stopifnot(is.numeric(df$prevalence), inherits(df$date, "Date"))


```


## Create Differenced Series - Differencing helps to make the time series stationary — good for LSTM training.
```{r}
# Anchor the original (level) series once
Series <- as.numeric(df$prevalence)


# Difference the prevalence
diffed <- diff(df$prevalence, differences = 1)


# Replace NA with 0 for first differenced value 
diffed <- c(0, diffed)

```


## Apply Your Lag Function
```{r}
lag_transform <- function(x, k = 1) {
  lagged = c(rep(NA, k), x[1:(length(x) - k)])
  DF = as.data.frame(cbind(lagged, x))
  colnames(DF) <- c(paste0("x-", k), "x")
  DF[is.na(DF)] <- 0
  return(DF)
}


supervised <- lag_transform(diffed, 1)
head(supervised)


```


## Split Dataset into Training & Testing Sets (70/30 Time-Series Split)

```{r}
## Split into train and test sets for time series

N <- nrow(supervised)              # Total observations
n <- round(N * 0.70, digits = 0)   # 70% split point

train <- supervised[1:n, ]         # First 70% as training
test  <- supervised[(n+1):N, ]     # Remaining 30% as testing


```

Unlike traditional machine learning where data can be randomly partitioned, time-series forecasting requires maintaining the temporal order of observations. Therefore, the first 70% of the lagged dataset was used for model training, while the remaining 30% was reserved for evaluating the LSTM forecast performance. This approach ensures that the model is trained on past values and tested only on future values, preventing data leakage and preserving the causal structure of the time series.



## Normalize the Data for LSTM
```{r}
## Function to scale data to desired feature range
scale_data <- function(train, test, feature_range = c(-1, 1)) {
  
  train_min <- min(train)
  train_max <- max(train)
  
  # Scale train & test using training min/max ONLY
  scaled_train <- ((train - train_min) / (train_max - train_min)) * 
                  (feature_range[2] - feature_range[1]) + feature_range[1]
                  
  scaled_test <- ((test - train_min) / (train_max - train_min)) * 
                 (feature_range[2] - feature_range[1]) + feature_range[1]
  
  # Return scaled sets and scaler values
  return(list(
    scaled_train = scaled_train,
    scaled_test  = scaled_test,
    scaler       = c(min = train_min, max = train_max)
  ))
}

## Apply scaling
Scaled <- scale_data(train, test, c(-1, 1))

# Extract scaled train & test values
x_train <- Scaled$scaled_train[,1]   # lagged values
y_train <- Scaled$scaled_train[,2]   # current values

x_test  <- Scaled$scaled_test[,1]
y_test  <- Scaled$scaled_test[,2]

```


LSTM models require scaling the data to a small numeric range. Since the sigmoid/tanh activation functions typically perform best within −1 to 1, we scale both training and testing sets to that range using only the training data min-max values. This avoids data leakage


## Correct Inverse-Scaling Function (Matching your [-1,1] scaling)

```{r}
## inverse transform function (scale back to original units)
invert_scaling <- function(scaled, scaler, feature_range = c(-1, 1)){
  
  min_val <- scaler[1]
  max_val <- scaler[2]
  
  ## unpack feature range
  fr_min <- feature_range[1]
  fr_max <- feature_range[2]
  
  # reverse scaling
  inverted <- (((scaled - fr_min) / (fr_max - fr_min)) * 
                (max_val - min_val)) + min_val
  
  return(inverted)
}


```


## Model

```{r}
# =========================
# Reshape to 3D [samples, timesteps, features]
# =========================
dim(x_train) <- c(length(x_train), 1, 1)
dim(x_test)  <- c(length(x_test),  1, 1)

X_shape2 <- dim(x_train)[2]   # timesteps = 1
X_shape3 <- dim(x_train)[3]   # features  = 1
batch_size <- 1               # must divide #samples; 1 is always safe
units <- 8                    # hidden size (tune as needed)

# =========================
# Build model (stateful LSTM)
# =========================
model <- keras_model_sequential() %>%
  layer_lstm(units,
             batch_input_shape = c(batch_size, X_shape2, X_shape3),
             stateful = TRUE,
             return_sequences = FALSE) %>%
  layer_dense(units = 1)

```


### compile
```{r}
model %>% compile(
  loss = "mse",
  optimizer = optimizer_adam(learning_rate = 0.001),
  metrics = c("mae")
)

summary(model)
```



### Fit the model

```{r}
epochs <- 50

for (e in 1:epochs) {
  model %>% fit(
    x = x_train, y = y_train,
    batch_size = batch_size,
    epochs = 1,
    shuffle = FALSE,
    verbose = 0
  )
  model %>% reset_states()
  if (e %% 10 == 0) cat("Epoch", e, "complete\n")
}

```

```{r}
# =========================
# Predict (train & test)
# =========================
train_pred_scaled <- model %>% predict(x_train, batch_size = batch_size)
model %>% reset_states()
test_pred_scaled  <- model %>% predict(x_test,  batch_size = batch_size)

# =========================
# Invert scaling to original units
# (We used feature_range = c(-1, 1) earlier)
# =========================
inv_scale <- function(z_scaled, scaler, feature_range = c(-1, 1)) {
  fr_min <- feature_range[1]; fr_max <- feature_range[2]
  train_min <- scaler["min"]; train_max <- scaler["max"]
  ((z_scaled - fr_min) / (fr_max - fr_min)) * (train_max - train_min) + train_min
}

y_train_hat <- inv_scale(train_pred_scaled[,1], Scaled$scaler, c(-1, 1))
y_test_hat  <- inv_scale(test_pred_scaled[,1],  Scaled$scaler, c(-1, 1))

y_train_true <- inv_scale(y_train, Scaled$scaler, c(-1, 1))
y_test_true  <- inv_scale(y_test,  Scaled$scaler, c(-1, 1))

```

```{r}
# =========================
# Evaluate
# =========================
rmse <- function(a, b) sqrt(mean((a - b)^2))
mae  <- function(a, b) mean(abs(a - b))

cat(sprintf("Train RMSE: %.4f | MAE: %.4f\n",
            rmse(y_train_true, y_train_hat),
            mae(y_train_true,  y_train_hat)))
cat(sprintf("Test  RMSE: %.4f | MAE: %.4f\n",
            rmse(y_test_true,  y_test_hat),
            mae(y_test_true,   y_test_hat)))

```


```{r echo=FALSE, fig.height=7, fig.width=10, message=FALSE, warning=FALSE, out.width="100%"}
# =========================
# Quick plots
# =========================
op <- par(mfrow = c(1,2))
plot(y_train_true, type = "l", main = "Train: True vs Pred",
     xlab = "Index", ylab = "Value")
lines(y_train_hat, col = 2)
legend("topleft", c("True","Pred"), lty = 1, col = c(1,2), bty = "n")

plot(y_test_true, type = "l", main = "Test: True vs Pred",
     xlab = "Index", ylab = "Value")
lines(y_test_hat, col = 2)
legend("topleft", c("True","Pred"), lty = 1, col = c(1,2), bty = "n")
par(op)




library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork) # for combining plots

# ---- Prepare training dataframe ----
df_train <- data.frame(
  Index = 1:length(y_train_true),
  Actual = y_train_true,
  Predicted = y_train_hat
) %>%
  pivot_longer(cols = c(Actual, Predicted), names_to = "Series", values_to = "Value")

p1 <- ggplot(df_train, aes(x = Index, y = Value, color = Series)) +
  geom_line(size = 0.5) +
  labs(
    title = "Train: Actual vs Predicted",
    x = "Index", y = "Prevalence"
  ) +
  scale_color_manual(values = c("Actual" = "blue", "Predicted" = "red")) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    plot.title = element_text(face = "bold")
  )

# ---- Prepare testing dataframe ----
df_test <- data.frame(
  Index = 1:length(y_test_true),
  Actual = y_test_true,
  Predicted = y_test_hat
) %>%
  pivot_longer(cols = c(Actual, Predicted), names_to = "Series", values_to = "Value")

p2 <- ggplot(df_test, aes(x = Index, y = Value, color = Series)) +
  geom_line(size = 0.5) +
  labs(
    title = "Test: Actual vs Predicted",
    x = "Index", y = "Prevalence"
  ) +
  scale_color_manual(values = c("Actual" = "blue", "Predicted" = "red")) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    plot.title = element_text(face = "bold")
  )

# ---- Combine side-by-side ----
p1 + p2


# Combine the two plots side by side
combined_plot <- p1 + p2

# Save as PNG
ggsave(
  filename = "HIV_Actual_vs_Predicted_Comparison.png",
  plot = combined_plot,
  width = 10,       # width in inches
  height = 5,       # height in inches
  dpi = 300         # resolution (suitable for publication)
)
```

```{r echo=FALSE, fig.height=7, fig.width=10, message=FALSE, warning=FALSE, out.width="100%"}
invert_scaling <- function(z_scaled, scaler, feature_range = c(-1, 1)) {
  fr_min <- feature_range[1]
  fr_max <- feature_range[2]
  train_min <- scaler["min"]
  train_max <- scaler["max"]
  
  ((z_scaled - fr_min) / (fr_max - fr_min)) * (train_max - train_min) + train_min
}




L <- length(x_test)
scaler <- Scaled$scaler
predictions <- numeric(L)

model %>% reset_states()

for(i in 1:L){
  X <- x_test[i]              # 1-step input
  dim(X) <- c(1,1,1)          # [samples, timestep, features]
  
  yhat_scaled <- model %>% predict(X, batch_size = batch_size)
  
  # invert scaling
  yhat <- invert_scaling(yhat_scaled, scaler, c(-1, 1))
  
  # invert differencing: add previous actual value
  yhat <- yhat + Series[n + i]
  
  predictions[i] <- yhat
}

actual_test <- Series[(n + 2):(n + L + 1)]
# Quick plot on original scale

# Quick plot on original scale
plot(actual_test, type="l", main="Test: Actual vs Reconstructed Forecast (Level)",
     xlab="Test Index", ylab="Prevalence")
lines(predictions, col=2)
legend("topleft", c("Actual","Forecast"), lty=1, col=c(1,2), bty="n")


# Ensure actual & predicted have same length
df_plot <- data.frame(
  Index = 1:length(actual_test),
  Actual = actual_test,
  Forecast = predictions
) %>% 
  tidyr::pivot_longer(cols = c(Actual, Forecast), names_to = "Series", values_to = "Value")

ggplot(df_plot, aes(x = Index, y = Value, color = Series)) +
  geom_line(size = 0.5) +
  labs(
    title = "LSTM Forecast vs Actual (Test Set)",
    subtitle = "Reconstructed HIV Prevalence on Original Scale",
    x = "Time Index (Test Period)",
    y = "Prevalence"
  ) +
  scale_color_manual(values = c("Actual" = "blue", "Forecast" = "red")) +
  theme_minimal(base_size = 12) +
  theme(
    legend.title = element_blank(),
    plot.title = element_text(face="bold")
  ) -> p

ggsave(
  filename = "HIV_Prevalence_Trajectory_2010_2021.png",
  plot = p,
  width = 8,        # width in inches
  height = 5,       # height in inches
  dpi = 300         # resolution (suitable for publication)
)

```


```{r echo=FALSE, fig.height=7, fig.width=10, message=FALSE, warning=FALSE, out.width="100%"}
# install.packages("DiagrammeR")  # if needed
library(DiagrammeR)

grViz("
digraph LSTMFlow {
  graph [rankdir=LR, fontsize=10, labelloc=t, labeljust=l]
  node  [shape=rectangle, style=filled, fillcolor=lightgrey, color=grey25, fontname=Helvetica, fontsize=10]
  edge  [color=grey35, arrowsize=0.7]

  A [label='Load data (xlsx) → df']
  B [label='Clean & tidy\\nfill(YEAR), Month→Month_num, date']
  C [label='Sort by date\\nselect(date, prevalence)']
  D [label='Difference series\\nprevalence → diffed (first NA→0)']
  E [label='Lag transform → supervised\\n(x-1, x)']
  F [label='Time-ordered split\\n70% train / 30% test']
  G [label='Scale to [-1,1]\\n(train min/max only)']
  H [label='Reshape to 3D\\n[samples, timesteps=1, features=1]']
  I [label='Build LSTM (stateful) + Dense(1)']
  J [label='Train (epochs loop)\\nshuffle=FALSE; reset_states()']
  K [label='Predict (scaled)\\ntrain & test']
  L [label='Invert scaling\\n→ original diff units']
  M [label='Invert differencing\\nŶ_level = Ŷ_diff + previous actual']
  N [label='Evaluate: RMSE/MAE']
  O [label='Visualize: lines\\nTrain/Test (ggplot)']

  A -> B -> C -> D -> E -> F -> G -> H -> I -> J -> K -> L -> M -> N -> O

  subgraph cluster_notes {
    label='Guards'
    style=dashed
    color=grey60
    G1 [label='No leakage:\\nuse train min/max', fillcolor=white]
    M1 [label='Use Series[n+i]\\nwhen reconstructing test', fillcolor=white]
  }
  G -> G1 [style=dashed]
  M -> M1 [style=dashed]
}
")

```

**Fig. 1: Data preprocessing pipeline (flowchart)**