---
title: "Deep Temporal Learning for HIV Forecasting: Modeling Epidemic Dynamics with Long Short-Term Memory Networks"
author: "Daniel Olofin"
date: "2025-10-26"
output: 
  pdf_document:
    latex_engine: xelatex
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading of packages

```{r}
library(readxl)
library(readr)
library(keras)
library(tensorflow)
library(viridis)
library(scales)
library(tidyr)
library(tidyverse)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(keras)
library(tensorflow)
library(purrr)
library(abind)
library(patchwork)
```


## Import Dataset

```{r}
HiV_data <- read_xlsx("Hivdata.xlsx")
```

## Data Cleaning, Feature Engineering and Basics Statistics

```{r echo=FALSE, fig.height=7, fig.width=10, message=FALSE, warning=FALSE, out.width="100%"}

# Rename for convenience
df <- HiV_data

# Handle Missing YEAR Values
df <- df %>%
  tidyr::fill(YEAR)

# Data Cleaning and Feature Engineering

df <- df %>%
  tidyr::fill(YEAR) %>%                                   # fill down missing YEARs
  mutate(
    Month = str_trim(Month),                       # Remove leading/trailing spaces
    Month = str_to_upper(Month),                   # Convert month names to uppercase
    Month_num = match(Month, toupper(month.abb)),  
    date = make_date(YEAR, Month_num, 1)           # Create full date variable (YYYY-MM-DD)
  )


# Summarize HIV Prevalence by Year
df_sum <- df %>%
  group_by(YEAR) %>%
  summarise(total = sum(prevalence, na.rm = TRUE)) 

# Identify the year with the highest total
max_year <- df_sum %>%
  filter(total == max(total)) %>%
  pull(YEAR)

```


## Data Visulization

```{r echo=FALSE, fig.height=7, fig.width=10, message=FALSE, warning=FALSE, out.width="100%"}

# Visualize annual totals 
p <- ggplot(df_sum, aes(x = factor(YEAR), y = total,
                   fill = ifelse(YEAR == max_year, "Highest", "Other"))) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = round(total, 2)), vjust = -0.3, size = 3) +
  scale_fill_manual(values = c("Highest" = "black", "Other" = "steelblue")) +
  labs(
  title = "Annual Aggregation of HIV Prevalence for Temporal Modeling (2010–2021)",
  x = "Year",
  y = NULL
) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none",
    plot.title = element_text(face = "bold")
  ) 

ggsave(
  filename = "p1.png",
  plot = p,
  width = 8,        # width in inches
  height = 5,       # height in inches
  dpi = 300         # resolution (suitable for publication)
)

# Monthly HIV Prevalence Plot 
p2 <- ggplot(df, aes(x = Month_num, y = prevalence, group = YEAR)) +
  geom_line(color="blue") +
  geom_point(color="red") +
  facet_wrap(~ YEAR, ncol = 3, scales = "free_y") + 
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1)) +

 labs(
  title = "Temporal Dynamics of Monthly HIV Prevalence (2010–2021)",
  x = "Month",
  y = "Prevalence"
)

ggsave(
  filename = "p2.png",
  plot = p2,
  width = 8,        # width in inches
  height = 5,       # height in inches
  dpi = 300         # resolution (suitable for publication)
)


# Plot: Temporal Trajectory of HIV Prevalence
p3 <- ggplot(df, aes(x = date, y = prevalence)) +
  geom_line(color = "blue") +
geom_point(color="red") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1)) +
  labs(
  title = "Temporal Trajectory of HIV Prevalence (2010–2021)",
  x = "Date",
  y = "Prevalence"
)

ggsave(
  filename = "p3.png",
  plot = p3,
  width = 8,        # width in inches
  height = 5,       # height in inches
  dpi = 300         # resolution (suitable for publication)
)
```


## Long Short-Term Memory Modeling

```{r}
# --- Reproducibility ---
set.seed(123)
tensorflow::tf$random$set_seed(123)

# --- 0) Data: ensure date is Date and sorted monthly ---
stopifnot(all(c("YEAR","Month","prevalence","Month_num","date") %in% names(df)))
df <- df %>%
  mutate(date = as.Date(date)) %>%
  arrange(date)

# Keep only modeling columns
df <- df %>% select(date, prevalence)


# --- Ensure date column and sort ---
df <- df %>%
  mutate(date = as.Date(date)) %>%
  arrange(date) %>%
  select(date, prevalence)

stopifnot(is.numeric(df$prevalence), inherits(df$date, "Date"))


```


## Create Differenced Series - Differencing helps to make the time series stationary — good for LSTM training.
```{r}
# Anchor the original (level) series once
Series <- as.numeric(df$prevalence)


# Difference the prevalence
diffed <- diff(df$prevalence, differences = 1)


# Replace NA with 0 for first differenced value 
diffed <- c(0, diffed)

```


## Apply Your Lag Function
```{r}
lag_transform <- function(x, k = 1) {
  lagged = c(rep(NA, k), x[1:(length(x) - k)])
  DF = as.data.frame(cbind(lagged, x))
  colnames(DF) <- c(paste0("x-", k), "x")
  DF[is.na(DF)] <- 0
  return(DF)
}


supervised <- lag_transform(diffed, 1)
head(supervised)


```


## Split Dataset into Training & Testing Sets (70/30 Time-Series Split)

```{r}
## Split into train and test sets for time series

N <- nrow(supervised)              # Total observations
n <- round(N * 0.70, digits = 0)   # 70% split point

train <- supervised[1:n, ]         # First 70% as training
test  <- supervised[(n+1):N, ]     # Remaining 30% as testing


```



## Normalize the Data for LSTM
```{r}
## Function to scale data to desired feature range
scale_data <- function(train, test, feature_range = c(-1, 1)) {
  
  train_min <- min(train)
  train_max <- max(train)
  
  # Scale train & test using training min/max ONLY
  scaled_train <- ((train - train_min) / (train_max - train_min)) * 
                  (feature_range[2] - feature_range[1]) + feature_range[1]
                  
  scaled_test <- ((test - train_min) / (train_max - train_min)) * 
                 (feature_range[2] - feature_range[1]) + feature_range[1]
  
  # Return scaled sets and scaler values
  return(list(
    scaled_train = scaled_train,
    scaled_test  = scaled_test,
    scaler       = c(min = train_min, max = train_max)
  ))
}

## Apply scaling
Scaled <- scale_data(train, test, c(-1, 1))

# Extract scaled train & test values
x_train <- Scaled$scaled_train[,1]   # lagged values
y_train <- Scaled$scaled_train[,2]   # current values

x_test  <- Scaled$scaled_test[,1]
y_test  <- Scaled$scaled_test[,2]

```




## Correct Inverse-Scaling Function (Matching your [-1,1] scaling)

```{r}
## inverse transform function (scale back to original units)
invert_scaling <- function(scaled, scaler, feature_range = c(-1, 1)){
  
  min_val <- scaler[1]
  max_val <- scaler[2]
  
  ## unpack feature range
  fr_min <- feature_range[1]
  fr_max <- feature_range[2]
  
  # reverse scaling
  inverted <- (((scaled - fr_min) / (fr_max - fr_min)) * 
                (max_val - min_val)) + min_val
  
  return(inverted)
}


```


## Modeling

```{r}
# =========================
# Reshape to 3D [samples, timesteps, features]
# =========================
dim(x_train) <- c(length(x_train), 1, 1)
dim(x_test)  <- c(length(x_test),  1, 1)

X_shape2 <- dim(x_train)[2]   # timesteps = 1
X_shape3 <- dim(x_train)[3]   # features  = 1
batch_size <- 1               # must divide #samples; 1 is always safe
units <- 8                    # hidden size (tune as needed)

# =========================
# Build model (stateful LSTM)
# =========================
model <- keras_model_sequential() %>%
  layer_lstm(units,
             batch_input_shape = c(batch_size, X_shape2, X_shape3),
             stateful = TRUE,
             return_sequences = FALSE) %>%
  layer_dense(units = 1)

```


### compile
```{r}
# --- Reproducibility ---
set.seed(123)
tensorflow::tf$random$set_seed(123)

model %>% compile(
  loss = "mse",
  optimizer = optimizer_adam(learning_rate = 0.001),
  metrics = c("mae")
)

summary(model)
```



### Fit the model

```{r}
# --- Reproducibility ---
set.seed(123)
tensorflow::tf$random$set_seed(123)

epochs <- 50

for (e in 1:epochs) {
  model %>% fit(
    x = x_train, y = y_train,
    batch_size = batch_size,
    epochs = 1,
    shuffle = FALSE,
    verbose = 0
  )
  model %>% reset_states()
  if (e %% 10 == 0) cat("Epoch", e, "complete\n")
}

```

##Prediction

```{r}
# --- Reproducibility ---
set.seed(123)
tensorflow::tf$random$set_seed(123)


# =========================
# Predict (train & test)
# =========================
train_pred_scaled <- model %>% predict(x_train, batch_size = batch_size)
model %>% reset_states()
test_pred_scaled  <- model %>% predict(x_test,  batch_size = batch_size)

# =========================
# Invert scaling to original units
# (We used feature_range = c(-1, 1) earlier)
# =========================
inv_scale <- function(z_scaled, scaler, feature_range = c(-1, 1)) {
  fr_min <- feature_range[1]; fr_max <- feature_range[2]
  train_min <- scaler["min"]; train_max <- scaler["max"]
  ((z_scaled - fr_min) / (fr_max - fr_min)) * (train_max - train_min) + train_min
}

y_train_hat <- inv_scale(train_pred_scaled[,1], Scaled$scaler, c(-1, 1))
y_test_hat  <- inv_scale(test_pred_scaled[,1],  Scaled$scaler, c(-1, 1))

y_train_true <- inv_scale(y_train, Scaled$scaler, c(-1, 1))
y_test_true  <- inv_scale(y_test,  Scaled$scaler, c(-1, 1))

```

## Matrices Evaluation
```{r}
# Evaluate
rmse <- function(a, b) sqrt(mean((a - b)^2))
mae  <- function(a, b) mean(abs(a - b))

cat(sprintf("Train RMSE: %.4f | MAE: %.4f\n",
            rmse(y_train_true, y_train_hat),
            mae(y_train_true,  y_train_hat)))
cat(sprintf("Test  RMSE: %.4f | MAE: %.4f\n",
            rmse(y_test_true,  y_test_hat),
            mae(y_test_true,   y_test_hat)))

```


```{r echo=FALSE, fig.height=7, fig.width=10, message=FALSE, warning=FALSE, out.width="100%"}

# --- Reproducibility ---
set.seed(123)
tensorflow::tf$random$set_seed(123)

# Quick plots

op <- par(mfrow = c(1,2))
plot(y_train_true, type = "l", main = "Train: True vs Pred",
     xlab = "Index", ylab = "Value")
lines(y_train_hat, col = 2)
legend("topleft", c("True","Pred"), lty = 1, col = c(1,2), bty = "n")

plot(y_test_true, type = "l", main = "Test: True vs Pred",
     xlab = "Index", ylab = "Value")
lines(y_test_hat, col = 2)
legend("topleft", c("True","Pred"), lty = 1, col = c(1,2), bty = "n")
par(op)




```


```{r echo=FALSE, fig.height=7, fig.width=10, message=FALSE, warning=FALSE, out.width="100%"}
# --- Reproducibility ---
set.seed(123)
tensorflow::tf$random$set_seed(123)

# ---- Prepare training dataframe ----
df_train <- data.frame(
  Index = 1:length(y_train_true),
  Actual = y_train_true,
  Predicted = y_train_hat
) %>%
  pivot_longer(cols = c(Actual, Predicted), names_to = "Series", values_to = "Value")

 ggplot(df_train, aes(x = Index, y = Value, color = Series)) +
  geom_line(size = 0.5) +
  labs(
    title = "Train: Actual vs Predicted",
    x = "Index", y = "Prevalence"
  ) +
  scale_color_manual(values = c("Actual" = "blue", "Predicted" = "red")) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    plot.title = element_text(face = "bold")
  )

# ---- Prepare testing dataframe ----
df_test <- data.frame(
  Index = 1:length(y_test_true),
  Actual = y_test_true,
  Predicted = y_test_hat
) %>%
  pivot_longer(cols = c(Actual, Predicted), names_to = "Series", values_to = "Value")

 ggplot(df_test, aes(x = Index, y = Value, color = Series)) +
  geom_line(size = 0.5) +
  labs(
    title = "Test: Actual vs Predicted",
    x = "Index", y = "Prevalence"
  ) +
  scale_color_manual(values = c("Actual" = "blue", "Predicted" = "red")) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    plot.title = element_text(face = "bold")
  )


```


```{r echo=FALSE, fig.height=7, fig.width=10, message=FALSE, warning=FALSE, out.width="100%"}
# --- Reproducibility ---
set.seed(123)
tensorflow::tf$random$set_seed(123)

invert_scaling <- function(z_scaled, scaler, feature_range = c(-1, 1)) {
  fr_min <- feature_range[1]
  fr_max <- feature_range[2]
  train_min <- scaler["min"]
  train_max <- scaler["max"]
  
  ((z_scaled - fr_min) / (fr_max - fr_min)) * (train_max - train_min) + train_min
}




L <- length(x_test)
scaler <- Scaled$scaler
predictions <- numeric(L)

model %>% reset_states()

for(i in 1:L){
  X <- x_test[i]              # 1-step input
  dim(X) <- c(1,1,1)          # [samples, timestep, features]
  
  yhat_scaled <- model %>% predict(X, batch_size = batch_size)
  
  # invert scaling
  yhat <- invert_scaling(yhat_scaled, scaler, c(-1, 1))
  
  # invert differencing: add previous actual value
  yhat <- yhat + Series[n + i]
  
  predictions[i] <- yhat
}

actual_test <- Series[(n + 2):(n + L + 1)]
# Quick plot on original scale

# Quick plot on original scale
plot(actual_test, type="l", main="Test: Actual vs Reconstructed Forecast (Level)",
     xlab="Test Index", ylab="Prevalence")
lines(predictions, col=2)
legend("topleft", c("Actual","Forecast"), lty=1, col=c(1,2), bty="n")


# Ensure actual & predicted have same length
df_plot <- data.frame(
  Index = 1:length(actual_test),
  Actual = actual_test,
  Forecast = predictions
) %>% 
  tidyr::pivot_longer(cols = c(Actual, Forecast), names_to = "Series", values_to = "Value")

ggplot(df_plot, aes(x = Index, y = Value, color = Series)) +
  geom_line(size = 0.5) +
  labs(
    title = "LSTM Forecast vs Actual (Test Set)",
    subtitle = "Reconstructed HIV Prevalence on Original Scale",
    x = "Time Index (Test Period)",
    y = "Prevalence"
  ) +
  scale_color_manual(values = c("Actual" = "blue", "Forecast" = "red")) +
  theme_minimal(base_size = 12) +
  theme(
    legend.title = element_blank(),
    plot.title = element_text(face="bold")
  ) 

ggsave(
  filename = "HIV_Prevalence_Trajectory_2010_2021.png",
  plot = p,
  width = 8,        # width in inches
  height = 5,       # height in inches
  dpi = 300         # resolution (suitable for publication)
)

```


```{r echo=FALSE, fig.height=7, fig.width=10, message=FALSE, warning=FALSE, out.width="100%"}
library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)

flow <- grViz("
digraph LSTMFlow {
  graph [rankdir=LR, fontsize=10, labelloc=t, labeljust=l]
  node  [shape=rectangle, style=filled, fillcolor=lightgrey, color=grey25, fontname=Helvetica, fontsize=10]
  edge  [color=grey35, arrowsize=0.7]

  A [label='Load data (xlsx) → df']
  B [label='Clean & tidy\\nfill(YEAR), Month→Month_num, date']
  C [label='Sort by date\\nselect(date, prevalence)']
  D [label='Difference series\\nprevalence → diffed (first NA→0)']
  E [label='Lag transform → supervised\\n(x-1, x)']
  F [label='Time-ordered split\\n70% train / 30% test']
  G [label='Scale to [-1,1]\\n(train min/max only)']
  H [label='Reshape to 3D\\n[samples, timesteps=1, features=1]']
  I [label='Build LSTM (stateful) + Dense(1)']
  J [label='Train (epochs loop)\\nshuffle=FALSE; reset_states()']
  K [label='Predict (scaled)\\ntrain & test']
  L [label='Invert scaling\\n→ original diff units']
  M [label='Invert differencing\\nŶ_level = Ŷ_diff + previous actual']
  N [label='Evaluate: RMSE/MAE']
  O [label='Visualize: lines\\nTrain/Test (ggplot)']
  A -> B -> C -> D -> E -> F -> G -> H -> I -> J -> K -> L -> M -> N -> O
}
")

svg <- export_svg(flow)

writeLines(svg, "lstm_flow.svg")

rsvg_png("lstm_flow.svg", "lstm_flow.png")
```

**Fig. 1: Data preprocessing pipeline (flowchart)**