---
title: "Statistical Method Assignment IV"
author: "Olofin Daniel Oluwafemi"
date: "2025-05-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 3
```{r message=FALSE, warning=FALSE}
#Load Packages
library(lubridate)
library(forecast)
library(tseries)
library(ggplot2)

# Loading Data
AssignmentIV <- read.delim("~/R for Statistics and Data Science (STAT-5521G)/AssignmentIV.txt")

```

```{r message=FALSE, warning=FALSE}
AssignmentIV$SP500 <- ts(AssignmentIV$SP500)

autoplot(AssignmentIV$SP500) +
  ggtitle("S&P 500 Time Series") +
  labs(x = "Time", y = "Index Value")+
  theme_minimal()
```


## 1. Does the data look stationary? Provide the R output. Do you need any transformation if you do provide an R output for it? (5pts).

**Null Hypothesis:** The time series is non-stationary. 
**Alternative Hypothesis:** The time series is stationary.

```{r message=FALSE, warning=FALSE}
# Plot ACF and PACF
acf(AssignmentIV$SP500, main = "ACF of SP500")
pacf(AssignmentIV$SP500, main = "PACF of SP500")

# Perform Augmented Dickey-Fuller Test
adf.test(AssignmentIV$SP500)
```
**Decision and Conclusion:** Since the p-value associated with the Dickey-Fuller test is greater than 0.05, we have all evidence to conclude that the data is non-stationary. Therefore, transformation and differencing of the data is inevitable. 

## 2. Provide an R output for test for stationarity. (5pts)

```{r message=FALSE, warning=FALSE}
#Transformation and Differencing
diff_sp <- diff(AssignmentIV$SP500)

# Test of Stationary
adf.test(diff_sp)
```

**Decision and Conclusion:** Since the p-value associated with the Dickey-Fuller test is lesser than 0.05 level of significance, we have all evidence to conclude that the data is stationary. Therefore, we can go ahead to modeling our data.


## 3. What is the order of the ARIMA model? Provide the R output of how did you find it.(10pts)

```{r message=FALSE, warning=FALSE}
model <- arima(diff_sp, order = c(1,1,3))
summary(model)
```


## 4. Report the theoretical model with the corresponding parameters. (5pts)

$$
\text{ARIMA}(1,1,3)
$$

This implies the model can be written in its theoretical form as:

$$
(1 - \phi_1 B)(1 - B)y_t = (1 + \theta_1 B + \theta_2 B^2 + \theta_3 B^3)\varepsilon_t
$$

Where:

- \( y_t \) is the S&P 500 index at time \( t \)
- \( B \) is the backshift operator
- \( \phi_1 \) is the autoregressive (AR) coefficient
- \( \theta_1, \theta_2, \theta_3 \) are the moving average (MA) coefficients
- \( \varepsilon_t \) is the white noise error term

So, the final theoretical ARIMA model is:

$$
(1 + 0.6169 B)(1 - B)y_t = (1 - 0.1644 B - 0.7313 B^2 - 0.0823 B^3)\varepsilon_t
$$


## 5. Provide an estimated values for the parameters, using R and the fitted model output. (10pts)
```{r message=FALSE, warning=FALSE}
model$coef
```


## 6. Provide a forecast plot for the model using R. (5 pts)

```{r message=FALSE, warning=FALSE}
autoplot(forecast(model))+
  labs(x = "Time", y = "Index Value")+
  theme_minimal()
```

