---
title: 'From Chaos to Clarity: Cleaning and Analyzing Dirty Restaurant Sales Data Using R'
author: "Moshood A. Garuba"
date: "2025-05-05"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background:

In the bustling world of restaurants, daily sales data is a goldmine—but only if it's clean and structured. Unfortunately, real-life data is rarely pristine. Missing values, inconsistencies, and incomplete entries are the norm rather than the exception. This project tackles one such challenge using a “dirty” restaurant sales dataset containing 17,534 transactions across starters, main dishes, desserts, drinks, and side dishes.

# The Problem:
The presence of dirt in data—such as missing categories, inconsistent entries, and incomplete transaction details—often makes analysis unreliable or impossible. Businesses that fail to clean and understand their data risk losing valuable insights that could improve operations, menu offerings, or customer satisfaction.

# Data Strategy:
- Data Importation
- Data Cleaning
- Exploratory Data Analysis
- Statistical Modeling and Inference

# Expected Outcome:
By the end of this project, the cleaned dataset will offer a reliable foundation for future analysis and modeling. Business insights—such as what items are most often paired, what times of day generate the most revenue, or what types of sales are prone to missing info—will be clearly visualized and discussed.


# Data Importation and Loading of Libraries
```{r message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)
library(lubridate)
library(readr)
library(janitor)
library(dplyr)
library(kableExtra)
library(naniar)
library(lubridate)
library(dlookr)
library(patchwork)
library(gt)
library(gtsummary)
library(report)

mydata <- read.csv("restaurant_sales_data.csv")

#checking the names of the variables
names(mydata)

#Checking the variable structure
str(mydata)

# Dimension
dim(mydata)

#Checking the first few observations
head(mydata) %>%  kable()

#Checking the last few observations
tail(mydata) %>%  kable()

```


# Data Cleaning:
```{r message=FALSE, warning=FALSE}
mydata %>%
  gg_miss_var()+
  labs(title = "Visaulization of the Missing variable")

sum(!complete.cases(mydata) == T)
```
**876 observations** are missing in the dataset. We need to check the pattern of missing and develop a strategy for adjusting the missingness either by omition or imputation method using some machine learning algorithms.


```{r message=FALSE, warning=FALSE}
plot_na_pareto(mydata)
```

**Observations:** From the Pareto na missing chart is show that the missing is fair and we can impute and fill the missing observations using machine learning algorithms called **Predictive mean matching**.

```{r message=FALSE, warning=FALSE}
# Step 1: Perform multiple imputation
imputed_data <- mice::mice(mydata, method = "pmm")

# Step 2: Extract the completed dataset (e.g., first imputation)
clean_data <- mice::complete(imputed_data, 1)

dim(clean_data)

gg_miss_var(clean_data)+
  labs(title = "Visaulization of the Missing variable after Imputation")
```


**Observation:** The only data problem is missing observation which has been corrected using **Predictive mean matching techniques** embended in `mice()` function and mice package.


# Data Analysis:
## Univariate analysis:

```{r message=FALSE, warning=FALSE}
clean_data %>% 
  group_by(Item) %>% 
  summarise(n = n()) %>% 
  ggplot(aes(y = Item, x = sort(n)))+
  geom_bar(stat = "identity", aes(fill = Item), show.legend = F)+
  theme_minimal()+
  labs(title = "Distributions of Items Ordered",
       y = "Items", x = "Frequency", caption = "Source: Kaggla")
```


```{r message=FALSE, warning=FALSE}
clean_data %>% 
  group_by(Category) %>% 
  summarise(n = n()) %>%
  mutate(Precentage = round(n/sum(n),3)*100) %>% kable()

```
**Obseveration:** The distribution of items across categories in the restaurant sales data is relatively balanced. Main Dishes have the highest count, making up 20.3% of all items sold (n = 3,551), closely followed by Starters (20.1%) and Desserts (19.9%). Drinks and Side Dishes account for 19.9% and 19.8% respectively. The differences among the categories are minimal, suggesting that customer demand is fairly evenly spread across all five item types, with a slight preference for main dishes.

```{r message=FALSE, warning=FALSE}
clean_data %>% 
  group_by(Category) %>% 
  summarise(n = n()) %>%
  mutate(Percentage = round(n/sum(n),3)*100) %>% 
  ggplot(aes(x = Category, y = Percentage ))+
  geom_bar(stat = "identity", aes(fill = Category), show.legend = F)+
  theme_minimal()+
    labs(title = "Distributions of Items Categories",
       y = "Categories", x = "Percentage", caption = "Source: Kaggla")
```



```{r message=FALSE, warning=FALSE}
clean_data %>% 
  summarytools::descr(Price, stat = "common")
```
**Observation:** The average price of items sold is **$6.60**, with a **standard deviation of $4.85**, indicating moderate variability in pricing. The **minimum price** observed is **$1.00**, while the **maximum** reaches **$20.00**, showing a wide range of item prices. The **median price** is **$5.00**, which is lower than the mean, suggesting a slight right-skew in the price distribution (i.e., a few higher-priced items pulling the average upward). All **17,534 observations** are valid, making up **100%** of the data, with no missing values for price.

```{r}
clean_data %>% 
  summarytools::descr(Quantity, stat = "common")
```

**Observation:** The average quantity of items ordered is **3.01**, with a **standard deviation of 1.42**, indicating a relatively consistent ordering pattern. The **minimum quantity** is **1**, and the **maximum** is **5**, suggesting that customers typically order between one and five items at a time. The **median quantity** is **3**, which aligns closely with the mean, indicating a fairly symmetrical distribution. All **17,534 observations** are valid, representing **100%** completeness for the `Quantity` variable.


```{r message=FALSE, warning=FALSE}
clean_data %>% 
  summarytools::descr(Order.Total, stat = "common")
```
**Observation:** The average order total is **$19.90**, with a **standard deviation of $18.83**, indicating considerable variability in how much customers spend per transaction. The **minimum order total** is **$1.00**, while the **maximum** reaches **$100.00**, showing a wide range of spending behavior. The **median** order total is **$15.00**, which is lower than the mean, suggesting a slight right-skew in the data due to some high-value orders. All **17,534 observations** are valid, reflecting **100%** data completeness for the `Order.Total` variable.

## Bivariate Analysis:

```{r message=FALSE, warning=FALSE}
clean_data %>% 
  group_by(Item) %>% 
  summarise(
    n = n(),
    `Total Quantity` = sum(Quantity, na.rm = TRUE)
  ) %>% 
  mutate(`Average Quantity` = round(`Total Quantity` / n, 2)) %>% 
  top_n(10, wt = `Total Quantity`) %>% kable()

```

```{r message=FALSE, warning=FALSE}
clean_data %>% 
  group_by(Item) %>% 
  summarise(
    n = n(),
    `Ordered Quantity` = sum(Order.Total, na.rm = TRUE)
  ) %>% 
  mutate(`Average Quantity` = round(`Ordered Quantity` / n, 2)) %>% 
  top_n(10, wt = `Ordered Quantity`) %>% kable()
```


```{r message=FALSE, warning=FALSE}
clean_data %>% 
  group_by(Category) %>% 
  summarise(
    n = n(),
    `Total Quantity` = sum(Quantity, na.rm = TRUE)
  ) %>% 
  mutate(`Average Quantity` = round(`Total Quantity` / n, 2)) %>% 
  top_n(10, wt = `Total Quantity`) %>% kable()
```


```{r}
clean_data %>% 
  group_by(Category) %>% 
summarise(
    n = n(),
    `Ordered Quantity` = sum(Order.Total, na.rm = TRUE)
  ) %>% 
  mutate(`Average Quantity` = round(`Ordered Quantity` / n, 2)) %>% 
  top_n(10, wt = `Ordered Quantity`) %>% kable()
```

# Statistical Modeling and Inference

```{r}
clean_data %>% 
  ggplot(aes(x = Price, y = Order.Total))+
  geom_point(aes(colour = Payment.Method))+
  labs(title = "Correlational Plot", y = "Total Order")+
  theme_minimal()
```

```{r}
model<-lm(Order.Total~Price, data = clean_data)

summary(model)
```

A linear regression model was fitted using ordinary least squares (OLS) to predict `Order.Total` based on `Price`. The model revealed a statistically significant and substantial relationship, explaining 62% of the variance in order totals (**R² = 0.62**, *F*(1, 17,532) = 28,132.81, *p* < .001). The intercept of the model was -0.23 (95% CI [-0.52, 0.07]), which was not statistically significant (*t* = -1.52, *p* = 0.128), indicating no meaningful prediction when price is zero. However, the effect of `Price` was highly significant and positive (**β = 3.05**, 95% CI [3.01, 3.08], *t* = 167.73, *p* < .001), with a standardized beta of **0.78** (95% CI [0.78, 0.79]), suggesting that higher prices are strongly associated with higher order totals. Overall, the model demonstrates that price is a strong and reliable predictor of total sales.