---
title: "Homework Two"
author: "Daniel Oluwafemi Olofin"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r include=FALSE}
#Making values not to read in scientific form
options(scipen = 999)
```

# a. Make a section with header “Visualization of trees data”. Under the section header, make a chunk that contains the command `plot(trees)`. Set the chunk options so that the code is not displayed, but the plot is:

```{r echo=FALSE}
plot(trees)
```

# b. Make a section with header “Area of a circle”. Under the section header, make a chunk that contains the command `area <- pi * r^2`. Set the chunk options so that the code is displayed, but not evaluated. (Grad student only). Insert the LaTeX code

```{r eval=FALSE}
area <- pi * r^2
```

Area of a Circle : $A = \pi r^2$

# c. Make a section with header “The great grey beast February”. Under the section header, make a chunk that calculates the number of seconds in the month of February (no leap year) and assignsthe value to a variable seconds. Set the chunk options so that the code is not displayed, but is evaluated. Write a line of text that states how many seconds are in February, making use of the value of the seconds variable
```{r echo=FALSE}
#Assigning the number of days in Feb to Num_of_days_in_Feb
Num_of_days_in_Feb <- 28
#Assigning the number of hours in a day to Num_of_hours_in_a_day
Num_of_hours_in_a_day <- 24
#Assigning the number of minutes in an hour to Num_of_minutes_in_an_hour
Num_of_minutes_in_an_hour <- 60
#Assigning the number of second in a minute to Num_of_seconds_in_a_minute
Num_of_seconds_in_a_Minute<- 60 

seconds <- Num_of_seconds_in_a_Minute * Num_of_minutes_in_an_hour * Num_of_hours_in_a_day * Num_of_days_in_Feb 
```
The number of seconds in February is **`r seconds`**.

# d. Make a section with header “Tasty dishes”. Consult the RMarkdown cheat sheet and see how to make an ordered list. Create an ordered list with at least three of your favorite food

1. Eba and Efo
2. Fried Rice
3. Dodo and Egg

