---
title: "Homework 4"
author: "Daniel Oluwafemi Olofin"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. This problem will have you create, modify, and expand a data frame.

## a. Create a small data frame describing a set of objects like I did with cardens. There should be at least four observations, and at least three variables, with at least one variable that is numeric and at least one that is a factor. The first column should be something usable as a unique identifier.

```{r}
# Create a data frame
fruit_df <- data.frame(
  ID = c("F001", "F002", "F003", "F004"),  # Unique identifier
  Name = factor(c("Apple", "Banana", "Cherry", "Date")),  # Factor variable
  Price = c(1.2, 0.5, 2.5, 3.0)  # Numeric variable
)

# View the data frame
print(fruit_df)

```

## b. Use logical indexing on the identifier column to select a row, and use a column/variable name to select a column, and use these to update one of the values in the data frame.

```{r}
# Update the price of Banana using logical indexing
fruit_df[fruit_df$ID == "F002", "Price"] <- 0.6

# Check the updated data frame
print(fruit_df)
```


## c. Use rbind() to add a new observation, and cbind() to add a new variable
```{r}
# Create a new row as a data frame
new_fruit <- data.frame(ID = "F005", Name = factor("Elderberry"), Price = 4.0)

# Define a new column (factor variable)
fruit_colors <- c("Red", "Yellow", "Red", "Brown", "Purple")

# Use rbind() to add the new observation
fruit_df <- rbind(fruit_df, new_fruit)

# Use cbind() to add the new column
fruit_df <- cbind(fruit_df, Color = factor(fruit_colors))

# View final data frame
print(fruit_df)

```

# 2. The code below will generate a small random matrix. Write code that will identify and remove the row with the smallest sum of absolute values.

```{r}
set.seed(123)
x <- matrix(sample(-5:5, 15, replace = TRUE), nrow = 5)

# Print the original matrix
print(x)

# Compute the sum of absolute values for each row
row_sums <- rowSums(abs(x))

# Identify the index of the row with the smallest sum
min_row_index <- which.min(row_sums)

# Remove the row with the smallest sum
x <- x[-min_row_index, , drop = FALSE]

# Print the updated matrix
print(x)
```

# 3. Identify and display the most frequent entry in the vector x. If there are ties, returning one of the values is sufficient. Hint: The table() function with argument decreasing = TRUE makes this much shorter than it might be otherwise.

```{r}
# Generate a random sample
x <- sample(1:5, 20, replace = TRUE)

# Print the vector
print(x)

# Create a frequency table, sorted in decreasing order
freq_table <- sort(table(x), decreasing = TRUE)

# Extract and display the most frequent value
most_frequent <- as.numeric(names(freq_table)[1])

print(most_frequent)

```

# 4. Highlight the following line and run it several times. Sometimes it returns a value, and sometimes it returns NULL. Explain why sometimes NULL is returned. Without changing the randomization aspect, find a way to ensure it never returns NULL.

```{r}
rownames(mtcars[ ,
sample(1:4, sample(1:4, size = 1))
])[[1]]
```

- **why is it null:** It returns `NULL` when no columns are selected from `mtcars`, which happens if `sample(1:4, size = 1)` randomly chooses `0`. In this case, `mtcars[ , selected_columns]` results in an empty data frame, and `rownames()` applied to an empty data frame returns an empty vector. Attempting to access `[[1]]` on an empty vector leads to `NULL`.

## solving it not to return NULL
```{r}
rownames(mtcars[ , sample(1:4, sample(1:4, size = 1)), drop = FALSE])[[1]]
```
# 5. Write a line of code that will remove any values from x that are not repeated.
Hint: My solution used %in%, as.numeric, names, table, >.

```{r}
x <- sample(c(-10:10), sample(20:40, 1), replace = TRUE)

# Remove values that appear only once
x <- x[x %in% as.numeric(names(table(x)[table(x) > 1]))]

# Print the updated x
print(x)

```

