---
title: "Statistics Method II Assignment"
author: "Moshood A Garuba"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question one:

## Codes:
```{r}
#Data Importation
mydata <- read.csv("~/R for Statistics and Data Science (STAT-5521G)/Stat_meth.txt", sep="")

#Linear Regression Model
lm_mdel<-lm(Minutes~Copiers, data = mydata)

#Summary of the LM result
summary(lm_mdel) -> summary_mdel
summary_mdel
```

## (a.)  What is the dependent variable in this case;
The dependent variable is: **total number of minutes**.

## (b.) What is the independent variable in this case;
The independent variable is: **number of copiers serviced**.

## (c.) Show what is the theoretical simple linear regression model;

 $$ Y = \beta_0 + \beta_1 X + \epsilon$$

where:

- \( Y \) = Total service time (minutes) (**dependent variable**),
- \( X \) = Number of copiers serviced (**independent variable**),
- \( \beta_0 \) = Intercept (baseline service time when no copiers are serviced),
- \( \beta_1 \) = Slope (additional time required per copier).

## (d.) Can you provide a least squares estimator for the parameter value \( \beta_1 \);

\( \beta_1 \) = `r summary_mdel$coefficients[2]`

## (e.) Can you provide a least squares estimator for the parameter value;

\( \beta_0 \) = `r summary_mdel$coefficients[1]`

## (f.) Show what is the fitted linear regression model;

  Total  service  time (minutes) = `r summary_mdel$coefficients[1]` + `r summary_mdel$coefficients[2]` (Number of copiers serviced) 
 
## (g.) Can you provide an interpretation for the slope \( \beta_1 \);
 
For one unit increase in number of copiers serviced, will lead to increase in the total serviced time on average by `r summary_mdel$coefficients[2]`. 
 
 
## (h.) Can you provide an interpretation for the intercept \( \beta_0 \):
 
 On average the total serviced time is `r summary_mdel$coefficients[1]` whenever the number of copiers serviced is 0.