---
title: "R Lab 4"
output: html_document
date: '2022-06-24'
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
rm(list = ls()) # clear our workspace. Always start with this
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
setwd("/Users/olofindo/Documents/Statsclinic")
```

# Correlation

We can assess the extent to which two variables are related to each
other. For instance, SBP at weeks 1 and 12.

### Example 6

```{r}
library(readr)
# bpfup <- read_csv("bpfup.csv ")

rds <- read_csv("residentdoctors.csv")
```

```{r}
#head(bpfup)

plot(rds$age,rds$mentalhealth_score)
```

```{r}
library(ggplot2)
#p <- ggplot(bpfup, aes(x = Week0, y=Week24)) + geom_point() + labs(y="SBP at week 24, mmHg", x="SBP at week 0, mmHg")
#p
```

```{r}
#cor(bpfup$Week0, bpfup$Week24, method = c("pearson"))

cor(rds$age,rds$mentalhealth_score, method = c("spearman"))
cor(rds$age,rds$mentalhealth_score, method = c("pearson"))

cor.test(rds$age,rds$mentalhealth_score, method = c("pearson"))

```

Note that there are other types of correlation - Kendall, Spearman.
Discussing them is beyond this class.

### Example 7

You asked 187 resident doctors about their hours worked and GHQ score, a
metric of mental health, and assess the extent to which both were
correlated.

```{r}
library(readr)
rds <- read_csv("residentdoctors.csv")
```

```{r}
head(rds)
```

```{r}
library(ggplot2)
p <- ggplot(rds, aes(x = hoursworked, y=mentalhealth_score)) + geom_point() + labs(y="Mental Health Score", x="Hours worked")
p
```

\*\*What do you notice? Is it possible that respondents were
approximating the number of hours they worked?

```{r}
cor(rds$mentalhealth_score, rds$hoursworked, method = c("pearson"))
```

\*\*What do you conclude?

# Simple Linear regression

Simple linear regression is also used to assess the extent to which two
variables may be related to one another. Linear regression is usually
preferable to correlation because: 1) Linear regression gives us an
effect measure which reveals the magnitude of the relationship 2) One of
the variables need not be continuous. It can be dichotomous or
categorical 3) Linear regression can include adjustment or control for
confounding.

### Example 8

Say you are interested in assessing whether there is a relationship
between the resident doctors' mental health and the number of hours
usually worked per week.

There are two common ways to program linear regression in R. 1)
lm([target variable] \~ [predictor variables], data = [data source]) 2)
glm([target variable] \~ [predictor variables], data = [data source],
family=gaussian(link = "identity"))

Here's how to implement the first approach

```{r}
model1 <- lm(mentalhealth_score ~ hoursworked, data = rds)
summary(model1)
```

\*\* How do you interpret your findings? Pay attention to: - the
estimate - the standard error - the p-value - the R-squared

Here's how to implement the second approach

```{r}
model2 <- glm(mentalhealth_score ~ hoursworked, data=rds, family=gaussian(link = "identity"))
summary(model2)

model2$fitted.values
round(mean(rds$mentalhealth_score - model2$fitted.values),1)
round(mean(model2$residuals),1)

```

###Evaluating regression models

We can assess Goodness of Fit

```{r}
#par(mfrow=c(2,2)) #tells R to arrange the next output in a 2 by 2 form
plot(model2)
```

### Binary predictor variable

The predictor/independent variable in a linear regression model can be a
binary or dichotomous. Let's illustrate this using the same recent
example.

Instead of using 'hoursworked' as a continuous variable, we could
examine whether the number of hours worked was greater than a
threshold - maybe 70 hours.

First, we create the binary variable

```{r}
library(dplyr) #which allows us use the pipe operator
rds$hoursgt70 <- rds %>%
  {ifelse(is.na(.$hoursworked), NA, #This checks for missing variables
         ifelse(.$hoursworked > 70, 1, 0))} 
              #Notice the logic: criteria, response if yes, response if no
```

This is the model with hours worked as a continuous variable

```{r}
model2 <- glm(mentalhealth_score ~ hoursworked, data=rds, family=gaussian(link = "identity"))
summary(model2)
```

Now, the model with hours worked as a binary variable

```{r}
model3 <- glm(mentalhealth_score ~ hoursgt70, data=rds, family=gaussian(link = "identity"))
summary(model3)
```

### Categorical Variables to test for trend

We can use categorical variables to test for linear trend.

First, we create a categorical variable

```{r}
rds$hourscat <- rds %>%
  {ifelse(is.na(.$hoursworked), NA,
         ifelse(.$hoursworked <= 40, 1,
                ifelse(.$hoursworked <= 60, 2,
                       ifelse(.$hoursworked <= 80, 3, 4))))}
```

```{r}
rds$hourscatx <- rds %>%
  {ifelse(is.na(.$hoursworked), NA,
         ifelse(.$hoursworked <= 40, "<=40 hours",
                ifelse(.$hoursworked <= 60, "41 - 60 hours",
                       ifelse(.$hoursworked <= 80, "61 - 80 hours", "81+ hours"))))}
```


Now we run the model. Note that the hourscat variable assumes that each
category is actually a numeric variable.

```{r}
model3 <- glm(mentalhealth_score ~ hourscat, data=rds, family=gaussian(link = "identity"))
summary(model3)
```

\*\*How do you interpret?

As an alternative, you can model the categorical variable, specifying to
R that you want this to be run as a categorical or factor variable.

```{r}
model3 <- glm(mentalhealth_score ~ as.factor(hourscatx), data=rds, family=gaussian(link = "identity"))
summary(model3)
```

### Indicator variables

Indicator or dummy variables are a special type of binary variables.
Essentially, you can convert a categorical variable to a number of
indicator variables, such that each level of the categorical variable is
represented by its own indicator variable.

We can create a dummy variable in R.

```{r}
rds$hourscat1 <- rds %>%
  {ifelse(is.na(.$hoursworked), NA,
         ifelse(.$hoursworked <= 40, 1,
                ifelse(.$hoursworked <= 60, 0,
                       ifelse(.$hoursworked <= 80, 0, 0))))}
rds$hourscat2 <- rds %>%
  {ifelse(is.na(.$hoursworked), NA,
         ifelse(.$hoursworked <= 40, 0,
                ifelse(.$hoursworked <= 60, 1,
                       ifelse(.$hoursworked <= 80, 0, 0))))}
rds$hourscat3 <- rds %>%
  {ifelse(is.na(.$hoursworked), NA,
         ifelse(.$hoursworked <= 40, 0,
                ifelse(.$hoursworked <= 60, 0,
                       ifelse(.$hoursworked <= 80, 1, 0))))}
rds$hourscat4 <- rds %>%
  {ifelse(is.na(.$hoursworked), NA,
         ifelse(.$hoursworked <= 40, 0,
                ifelse(.$hoursworked <= 60, 0,
                       ifelse(.$hoursworked <= 80, 0, 1))))}
```

When we have multiple parameters in a model, we use the F statistic to
decide whether to look at the individual beta estimates. This is the
Omnibus Test.

```{r}
model4 <- lm(mentalhealth_score ~ hourscat1 + hourscat3 + hourscat4, data = rds)
summary(model4)
```

\*\* What do you conclude?

# Multiple Linear regression

Sometimes, we can have more than one predictor variable or covariate. A
number of reasons why we may have multiple predictors 1) to improve
prediction 2) to adjust for confounding

Here's the model we saw earlier

```{r}
model4 <- lm(mentalhealth_score ~ hourscat1 + hourscat3 + hourscat4, data = rds)
summary(model4)
```

Say you believe that gender is important to consider. So, you include it
in the model. This is now a multiple linear regression, or a
multivariable linear regression.

```{r}
model5 <- lm(mentalhealth_score ~ hourscat1 + hourscat3 + hourscat4 + female, data = rds)
summary(model5)
```

Take a look at the R-squared: without the female variable vs with the
female variable. Simply adding one variable leads to a higher R-squared,
whether that new variable improves your model or not.

Hence, we depend on the adjusted R-squared. The adj. R-squared is
adjusted for the number of variables.

Let's try a different covariate - the number of children the resident
has, a count variable

```{r}
model6 <- lm(mentalhealth_score ~ hourscat2 + hourscat3 + hourscat4 + children, data = rds)
summary(model6)
```

We could also use the categorical form of the variable. Here, you have
to be careful to not fall into the trap of modeling your categorical
variable as a continuous or numeric variable as shown here.

```{r}
rds$chlcat <- rds %>%
  {ifelse(is.na(.$children), NA,
         ifelse(.$children %in% 0, 0,
                ifelse(.$children %in% 1, 1,
                       ifelse(.$children %in% 2, 2, 3))))}
```

```{r}
model7 <- lm(mentalhealth_score ~ hourscat2 + hourscat3 + hourscat4 + chlcat, data = rds)
summary(model7)
```

Instead, this is one way to do it

```{r}
model7 <- lm(mentalhealth_score ~ hourscat2 + hourscat3 + hourscat4 + as.factor(chlcat), data = rds)
summary(model7)
```

You can compare models 4, without the number of children variable and 7,
with the number of children

```{r}
AIC(model4, model7)

AIC(model4,model7)
```

Here, we see that model 7 has the smaller AIC. We sometimes use a rule
of thumb of **10**. Need to confirm

If you also compare the adjusted R-squared for models 4 and 7, you would
see that the R-squared is larger for model 7 suggesting improved
prediction.

You can also add more variables. There are often a number of ways to
decide what variables to include in the model.

```{r}
model8 <- lm(mentalhealth_score ~ hourscat2 + hourscat3 + hourscat4 + as.factor(chlcat) + as.factor(agecat), data = rds)
summary(model8)
```

```{r}
AIC(model7, model8)
```

### Interaction terms

Sometimes, we may want to test to see if there is effect modification or
some interaction. Interaction refers to a different magnitude of
relationship between two variables depending on the levels of a 3rd
variable.

Here is the model we just looked at earlier

```{r}
model8 <- lm(mentalhealth_score ~ as.factor(hourscat) + as.factor(chlcat) + as.factor(agecat), data = rds)
summary(model8)
```

Here is the same model with an interaction term for sex

```{r}
model9 <- lm(mentalhealth_score ~ as.factor(hourscat) + as.factor(chlcat) * as.factor(agecat), data = rds)
summary(model9)
```

```{r}
AIC(model8, model9)
```

\*What do you conclude? Is model 8 better than model 9? If the models
are not significantly different, which one is more parsimonious?

# Non-linearity

We can assess the nonlinearity of the relationship of continuous
variables.

```{r}
model10 <- lm(mentalhealth_score ~ hoursworked, data = rds)
summary(model10)
```

```{r}
rds$hours_sqd <- rds$hoursworked*rds$hoursworked
model11 <- lm(mentalhealth_score ~ hoursworked + hours_sqd, data = rds)
summary(model11)
```

We can use the Akaike Information Criterion (AIC) to choose the better
model. The model with the smaller AIC is the better model, if the
difference is \>10. If there is no meaningful difference in the AIC, we
would opt for the simpler or more parsimonious model

```{r}
AIC(model10, model11)
```

If the p-value for the quadratic variable is not significant, then,
there is no need to examine the cubed variable. Lets pretend for
didactic purposes and go ahead to show you what the higher order
analysis will look like.

```{r}
rds$hours_cubed <- rds$hoursworked*rds$hoursworked*rds$hoursworked
model12 <- lm(mentalhealth_score ~ hoursworked + hours_sqd + hours_cubed, data = rds)
summary(model12)
```

```{r}
AIC(model10, model12)
```
