---
title: "Assigment 8"
author: "Moshood Abiola"
date: "2025-09-22"
output:
  pdf_document:
    latex_engine: xelatex
always_allow_html: true
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1) This problem will step you through making a prediction using more than one predictor variable

### a) Modify Example 8.7 (continued) to predict the price of a car with 200 horsepower, a weight of 2705, and a length of 177. Use **k = 3**. You should get 19.533.


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(MASS)    
library(caret)   

# Keep only complete cases of predictors and response
car_data <- subset(Cars93, select = c(Price, Horsepower, Weight, Length))
car_data <- na.omit(car_data)

# Query observation
query_pt <- data.frame(Horsepower = 200, Weight = 2705, Length = 177)

# 1a) KNN regression on raw features
set.seed(123)
knn_raw <- knnreg(Price ~ Horsepower + Weight + Length, data = car_data, k = 3)
pred_raw_knn <- predict(knn_raw, query_pt)
pred_raw_knn
```

### b.) Unfortunately this isn’t what I obtained when I tested it with my homemade code. I wondered if this is because knnreg() isn’t standardizing the predictor variables, so I tested it this way: I made a subset of the Cars93 dataset only containing the predictor variables, scaled it and turned it back into a data frame, then added the Price variable. I fit a model used this scaled dataset. Then I scaled the observation I wanted to predict for by the predictor variables and coerced it to a data fame, and finally made a prediction. It was 17.4. Obtain this same prediction.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# 1b) Standardize predictors (center + scale), then KNN
X_train <- car_data[, c("Horsepower","Weight","Length")]

scaler <- preProcess(X_train, method = c("center","scale"))

X_train_scaled <- predict(scaler, X_train)
car_data_scaled <- data.frame(X_train_scaled, Price = car_data$Price)

set.seed(123)
knn_scaled <- knnreg(Price ~ ., data = car_data_scaled, k = 3)

query_scaled <- predict(scaler, query_pt)
pred_scaled_knn <- predict(knn_scaled, query_scaled)
pred_scaled_knn              
```


### 2.) (Grad student only) Write a function that implements Algorithm 8.6. Use it to solve problem 1) and ensure your answer is the same as in 1b).


```{r echo=FALSE, message=FALSE, warning=FALSE}
manual_knn <- function(X, y, x_new, k = 3, scale_it = FALSE) {
  stopifnot(nrow(X) == length(y),
            ncol(X) == ncol(x_new),
            k >= 1, k <= nrow(X))
  
  X_mat <- as.matrix(X)
  q_mat <- as.matrix(x_new)
  
  if (scale_it) {
    mu_vals <- colMeans(X_mat)
    sd_vals <- apply(X_mat, 2, sd)
    sd_vals[sd_vals == 0] <- 1
    X_mat <- scale(X_mat, center = mu_vals, scale = sd_vals)
    q_mat <- scale(q_mat,  center = mu_vals, scale = sd_vals)
  }
  
  # Euclidean distances
  dist_vec <- sqrt(rowSums((X_mat - matrix(q_mat, nrow(X_mat), ncol(X_mat), byrow = TRUE))^2))
  
  # k nearest indices
  nearest_idx <- order(dist_vec)[1:k]
  
  # Mean of target values
  mean(y[nearest_idx])
}

```


#### Reproduce 1a with manual KNN (no standardization)

```{r echo=FALSE, message=FALSE, warning=FALSE}
pred_raw_manual <- manual_knn(
  X = car_data[, c("Horsepower","Weight","Length")],
  y = car_data$Price,
  x_new = query_pt,
  k = 3,
  scale_it = FALSE
)
pred_raw_manual 
```


#### Reproduce 1b with manual KNN (with standardization)

```{r echo=FALSE, message=FALSE, warning=FALSE}
pred_scaled_manual <- manual_knn(
  X = car_data[, c("Horsepower","Weight","Length")],
  y = car_data$Price,
  x_new = query_pt,
  k = 3,
  scale_it = TRUE
)
pred_scaled_manual 

```

