---
title: "Homework 9"
author: "Moshood Abiola"
date: "2025-09-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## (1.) This problem will step you through fitting a regression tree model and making a prediction

### (a.) Use the Cars93 dataset in the MASS library. Create a tree for predicting Price from
Horsepower, Weight, Length, whether a car has AirBags, and the Origin of production.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(MASS)
library(rpart)
library(rpart.plot)

# Build regression tree
fit_model <- rpart(Price ~ Horsepower + Weight + Length + AirBags + Origin, 
             data = Cars93, method = "anova")

# Visualize
rpart.plot(fit_model)

```


### (b.) Predict the price of a car with 200 horsepower, weighs 2705 pounds, is 177 inches long, has
driver and passenger airbags, and was not made in the US.

```{r echo=FALSE, message=FALSE, warning=FALSE}
newcar_data <- data.frame(Horsepower = 200,
                     Weight = 2705,
                     Length = 177,
                     AirBags = "Driver & Passenger",
                     Origin = "non-USA")

predict(fit_model, newdata = newcar_data)

```


### (c.) Use `plotcp()` to decide on an optimal cp value. You have a little flexibility in your choice, but explain why you picked the one you did. Use `prune()` to apply the chosen cp, and plot the pruned tree.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Pick cp
printcp(fit_model)   # shows table
plotcp(fit_model)    # visualize

# Suppose cp = 0.02 looks optimal
pruned <- prune(fit_model, cp = 0.02)

# Plot pruned tree
rpart.plot(pruned)

```

## 2. (Grad student only) I wrote most of a function to implement Algorithm 9.2, but left the innermost parts out. I also provided two functions to aid the construction and retrieval of all subsets of levels of a categorical variable.

### (a.) Complete the function 

```{r echo=TRUE, message=FALSE, warning=FALSE}

# I fix the first level on the LEFT to avoid duplicates like {A}|{B,C} and {B,C}|{A}.
level_partitions <- function(levels_vec) {
  L <- sort(unique(levels_vec)); k <- length(L)
  if (k <= 1) return(list())
  first <- L[1]; others <- L[-1]
  out <- list()
  # iterate all subsets of 'others'; exclude empty (only {first}) and full (all levels)
  for (m in 1:(2^(k-1) - 1)) {
    pick <- as.logical(intToBits(m))[seq_along(others)]
    left <- c(first, others[pick])
    if (length(left) < k) out[[length(out) + 1]] <- left
  }
  out
}

# Weighted mean and weighted SSE
wmean <- function(x, w) sum(w * x) / sum(w)
wsse  <- function(x, w) { mu <- wmean(x, w); sum(w * (x - mu)^2) }


best_split_alg9_2 <- function(X, y,
                              weights  = NULL,
                              minsplit = 2,     # small defaults for classroom toy data
                              minbucket = NULL,
                              cp = 0) {
  n <- length(y)
  if (is.null(weights)) weights <- rep(1, n)
  if (n < minsplit) return(NULL)
  if (is.null(minbucket)) minbucket <- max(1, floor(minsplit / 3))

  # 1) baseline goodness = SSE of parent node
  sse_parent <- wsse(y, weights)

  # initialize
  best <- list(goodness = Inf, split_var = NA, split_val = NA,
               left_idx = integer(0), right_idx = integer(0),
               improvement = 0, type = NA)

  # 2) loop over columns of X
  for (j in seq_len(ncol(X))) {
    xj <- X[[j]]

    # (a) numeric
    if (is.numeric(xj)) {
      ux <- sort(unique(xj))
      if (length(ux) > 1) {
        mids <- (ux[-1] + ux[-length(ux)]) / 2
        for (thr in mids) {
          L <- which(xj <= thr); R <- which(xj > thr)
          if (length(L) < minbucket || length(R) < minbucket) next
          sse_LR <- wsse(y[L], weights[L]) + wsse(y[R], weights[R])
          improve <- sse_parent - sse_LR
          if (sse_LR < best$goodness && improve >= cp * sse_parent) {
            best <- list(goodness = sse_LR, split_var = names(X)[j],
                         split_val = thr, left_idx = L, right_idx = R,
                         improvement = improve, type = "numeric")
          }
        }
      }

    # (b) categorical
    } else {
      levs <- unique(xj)
      if (length(levs) > 1) {
        parts <- level_partitions(levs)
        for (left_levels in parts) {
          L <- which(xj %in% left_levels); R <- setdiff(seq_len(n), L)
          if (length(L) < minbucket || length(R) < minbucket) next
          sse_LR <- wsse(y[L], weights[L]) + wsse(y[R], weights[R])
          improve <- sse_parent - sse_LR
          if (sse_LR < best$goodness && improve >= cp * sse_parent) {
            best <- list(goodness = sse_LR, split_var = names(X)[j],
                         split_val = left_levels, left_idx = L, right_idx = R,
                         improvement = improve, type = "categorical")
          }
        }
      }
    }
  }

  if (is.infinite(best$goodness)) return(NULL)
  best
}
```


### 2.b Test the completed function on the toy dataset in Example 9.1 and ensure it returns the same
answer we obtained by hand
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Create the toy dataset
X_1 <- c(2, 8, 4, 3, 6)
X_2 <- factor(c("Red", "Green", "Red", "Blue", "Green"))
Y   <- c(3, 10, 5, 3, 6)
toyX <- data.frame(X_1 = X_1, X_2 = X_2)

# Run Algorithm 9.2
ans <- best_split_alg9_2(toyX, Y, minsplit = 2, minbucket = 1, cp = 0)

# Report
cat("split_var :", ans$split_var, "\n")
cat("split_val :", if (is.numeric(ans$split_val)) ans$split_val else paste(ans$split_val, collapse=", "), "\n")
cat("goodness  :", round(ans$goodness, 4), "\n")
cat("left idx  :", paste(ans$left_idx, collapse = " "), "\n")
cat("right idx :", paste(ans$right_idx, collapse = " "), "\n")

```

