---
title: "Home work 11"
author: "Moshood A Garuba"
date: "2025-10-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

Use the `Cars93` dataset in the MASS library. Create a linear model for predicting Price from Horsepower, Weight, Length, whether a car has AirBags, and the Origin of production. Use it to predict the price of a car with 200 horsepower, weighs 2705 pounds, is 177 inches long, has driver and passenger airbags, and was not made in the USA.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(MASS)
df <- Cars93

lm_model <- lm(Price ~ Horsepower + Weight + Length + AirBags + Origin, data = df)

summary(lm_model)

```

## Question 2

(Grad student only) Obtain regression coefficients for the model in problem 1), but using gradient descent. They should be practically equivalent. Use the grad_descent() and lin_reg() functions I have provided.

Hint: To deal with the categorical predictors, you need to make a binary column for each level that is not the default. The value should be 1 if the observation is in that category and 0 otherwise. So you can check your work, here is what the first several rows of the data matrix should look like. The tools I used are `cbind()`, `==`comparison between a dataframe column referenced by $ and a string, and
the fact that logicals can be implicitly coerced to numeric.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Check factor levels (for reference)
levels(Cars93$AirBags)
levels(Cars93$Origin)


new_car <- data.frame(
  Horsepower = 200,
  Weight = 2705,
  Length = 177,
  AirBags = "Driver & Passenger",
  Origin = "non-USA"
)



predicted_price <- predict(lm_model, newdata = new_car)
predicted_price


```



```{r echo=FALSE, message=FALSE, warning=FALSE}
X <- cbind(
  1,  # Intercept column
  Cars93$Horsepower,
  Cars93$Weight,
  Cars93$Length,
  as.numeric(Cars93$AirBags == "Driver only"),
  as.numeric(Cars93$AirBags == "None"),
  as.numeric(Cars93$Origin == "non-USA")
)

Y <- Cars93$Price

# Assign column names (for clarity only)
colnames(X) <- c("Intercept", "Horsepower", "Weight", "Length",
                 "AirBags_DriverOnly", "AirBags_None", "Origin_nonUSA")



#  Define gradient descent and linear regression functions

grad_descent <- function(func, grad_func, eta, threshold, x0,
                         show_plot = FALSE, X, Y) {
  x <- x0
  keep_going <- TRUE
  iter <- 0
  
  while (keep_going) {
    grad <- grad_func(x, X, Y)
    new_x <- x - eta * grad
    iter <- iter + 1

    if (sqrt(sum(grad^2)) < threshold) {
      keep_going <- FALSE
    }
    x <- new_x
  }
  
  cat("Converged after", iter, "iterations.\n")
  cat("Local minimum of", func(x, X, Y), "found at:\n")
  print(x)
  
  return(list(min_location = x, min_value = func(x, X, Y)))
}


lin_reg <- function(X, Y, eta = .001, threshold = .001,
                    beta0 = rep(0, ncol(X)), show_plot = FALSE) {
  func <- function(beta, X, Y) {
    sum((Y - X %*% beta) ^ 2)
  }
  grad_func <- function(beta, X, Y) {
    -2 * t(X) %*% (Y - X %*% beta)
  }
  
  # Standardize X (except intercept) and Y
  Xsc <- cbind(1, scale(X[, -1]))
  Ysc <- scale(Y)
  
  # Run gradient descent
  gd_out <- grad_descent(func, grad_func, eta, threshold, beta0,
                         show_plot = show_plot, X = Xsc, Y = Ysc)
  
  beta_sc <- gd_out$min_location
  
  # Convert back to unstandardized coefficients
  beta <- c(
    sum(-beta_sc[-1] * colMeans(X)[-1] * sd(Y) / apply(X, 2, sd)[-1]) + mean(Y),
    beta_sc[-1] * sd(Y) / apply(X, 2, sd)[-1]
  )
  
  cat("\nUnstandardized regression coefficients:\n")
  print(beta)
  
  return(beta)
}


# Run gradient descent regression
beta_gd <- lin_reg(X, Y, eta = 0.001, threshold = 0.001)


# Compare with ordinary least squares
model_lm <- lm(Price ~ Horsepower + Weight + Length + AirBags + Origin, data = Cars93)
summary(model_lm)$coefficients
```

