---
title: "Applied Regression Ass 1"
author: "Daniel Olofin"
date: "2025-09-04"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## (a) Why is the model linear and non-linear?

We model the circle area \(A\) as
\[
A=\beta_0+\beta_1 x^2+\varepsilon.
\]
- **Linear (in parameters):** The mean function is an affine combination of \(\beta_0,\beta_1\); i.e., linear in the unknown coefficients.  
- **Non-linear (in predictor):** The regressor is \(x^2\), a non-linear transform of \(x\).

## (b) Interpretations of \(\beta_0\) and \(\beta_1\)

- \(\beta_0\): Expected area when \(x=0\) (intercept). For true circles, we expect \(\beta_0\approx 0\).  
- \(\beta_1\): Expected change in area **per unit increase in \(x^2\)**. Geometrically, the true value is \(\pi\) since \(A=\pi x^2\).

## (c) Why include the error term \(\varepsilon\)? Give two sources.

We include \(\varepsilon\) to capture randomness and model imperfections, e.g.:
- **Measurement error** in radius or area.
- **Omitted factors/model misspecification** (e.g., deformation, thickness), rounding, and sampling variability.

## (d) Sample version for \(n\) observations

For \(i=1,\dots,n\),
\[
A_i=\beta_0+\beta_1 x_i^2+\varepsilon_i,
\qquad \mathbb{E}[\varepsilon_i]=0,\ \ \mathrm{Var}(\varepsilon_i)=\sigma^2.
\]

## (e) Vector/matrix form and dimensions

Let
\[
\mathbf{A}=
\begin{bmatrix}
A_1\\ \vdots\\ A_n
\end{bmatrix}_{n\times 1},\quad
\mathbf{X}=
\begin{bmatrix}
1 & x_1^2\\
\vdots & \vdots\\
1 & x_n^2
\end{bmatrix}_{n\times 2},\quad
\boldsymbol{\beta}=
\begin{bmatrix}
\beta_0\\ \beta_1
\end{bmatrix}_{2\times 1},\quad
\boldsymbol{\varepsilon}=
\begin{bmatrix}
\varepsilon_1\\ \vdots\\ \varepsilon_n
\end{bmatrix}_{n\times 1}.
\]
Then
\[
\mathbf{A}=\mathbf{X}\boldsymbol{\beta}+\boldsymbol{\varepsilon}.
\]
**Dimensions:** \(\mathbf{A}\in\mathbb{R}^{n\times 1}\), \(\mathbf{X}\in\mathbb{R}^{n\times 2}\), \(\boldsymbol{\beta}\in\mathbb{R}^{2\times 1}\), \(\boldsymbol{\varepsilon}\in\mathbb{R}^{n\times 1}\).

## (f) Use to fit the model to the data (see next slide)

```{r echo=FALSE, message=FALSE, warning=FALSE}
radius <- c(1,2,3,4,5,6)
radius_squared <- radius^2
area <- c(3.5,12.9,27.8,52.1,79.0,113.5)

our_data <- data.frame(radius,radius_squared, area)
```

### (i) Graph a scatterplot for A against x^2

```{r echo=FALSE, message=FALSE, warning=FALSE}
plot(y = our_data$area, x = our_data$radius_squared,
     col = "blue", main = "Scatterplot for A against radius square", ylab = "Area", xlab = "Radius square")
```

### (iii) Run linear regression analysis in R for the simple linear regression model. Copy the output and write the equation for the fitted model.

```{r echo=FALSE, message=FALSE, warning=FALSE}
fit <- lm(area~radius_squared, data = our_data)

summary(fit)

```

\[
\hat{A} = \texttt{0.34068} + \texttt{3.15116} \cdot (\text{radius}^2)
\]

### (iii) Using R, graph the fitted model on the scatterplot  in (i)

```{r echo=FALSE, message=FALSE, warning=FALSE}
plot(y = our_data$area, x = our_data$radius_squared,
     col = "blue", main = "Scatterplot for A against radius square", ylab = "Area", xlab = "Radius square")

abline(lm(area~radius_squared, data = our_data), col ="red")
```


## (g) Compare simple vs. multiple regression

We compare  
- **Simple model:** \(A=\beta_0+\beta_1 x^2+\varepsilon\)  
- **Multiple model:** \(A=\beta_0+\beta_1 x+\beta_2 x^2+\varepsilon\)  
with \(x=\texttt{radius}\), \(x^2=\texttt{radius\_squared}\).

### (i) Why is \(A=\beta_0+\beta_1 x+\beta_2 x^2+\varepsilon\) linear **and** multiple?
- **Linear (in parameters):** \(A\) is an affine function of the unknowns \(\beta_0,\beta_1,\beta_2\) (no products/powers of \(\beta\)’s).  
- **Multiple:** it uses **two regressors** (\(x\) and \(x^2\)) for one response, hence a multiple regression.

### (ii) Sample (observation-level) model
For \(i=1,\dots,n\):  
\[
A_i=\beta_0+\beta_1 x_i+\beta_2 x_i^2+\varepsilon_i,\qquad
\mathbb E[\varepsilon_i]=0,\ \text{Var}(\varepsilon_i)=\sigma^2.
\]

### (iii) Vector/matrix form and dimensions
Let  
\[
\mathbf A=
\begin{bmatrix}A_1\\ \vdots\\ A_n\end{bmatrix}_{n\times1},\quad
\mathbf X=
\begin{bmatrix}
1 & x_1 & x_1^2\\
\vdots & \vdots & \vdots\\
1 & x_n & x_n^2
\end{bmatrix}_{n\times3},\quad
\boldsymbol\beta=
\begin{bmatrix}\beta_0\\ \beta_1\\ \beta_2\end{bmatrix}_{3\times1},\quad
\boldsymbol\varepsilon=
\begin{bmatrix}\varepsilon_1\\ \vdots\\ \varepsilon_n\end{bmatrix}_{n\times1}.
\]
Then \(\mathbf A=\mathbf X\boldsymbol\beta+\boldsymbol\varepsilon\).  
**Dimensions:** \(\mathbf A\in\mathbb R^{n\times1}\), \(\mathbf X\in\mathbb R^{n\times3}\), \(\boldsymbol\beta\in\mathbb R^{3\times1}\), \(\boldsymbol\varepsilon\in\mathbb R^{n\times1}\).


### (iv) Graph a scatterplot for A against x

```{r echo=FALSE, message=FALSE, warning=FALSE}
plot(y = our_data$area, x = our_data$radius,
     col = "blue", main = "Scatterplot for A against radius ", ylab = "Area", xlab = "Radius")
```

### (v) Run linear regression analysis in R for the simple linear regression model. Copy the output and write the equation for the fitted model.

```{r echo=FALSE, message=FALSE, warning=FALSE}
fit_mul <- lm(area~radius+radius_squared, data = our_data)

summary(fit_mul)

```

\[
\hat{A} = -0.2100 + 0.3868 \cdot (\text{radius}) + 3.0982 \cdot (\text{radius}^2)
\]

### (vi) Using R, graph the fitted model on the scatterplot  in (i)

```{r echo=FALSE, message=FALSE, warning=FALSE}
# scatter
plot(x = our_data$radius, y = our_data$area,
     col = "blue", main = "Scatterplot for Area vs Radius",
     xlab = "Radius", ylab = "Area", pch = 16)

# grid of x values over the observed range
xx <- seq(min(our_data$radius, na.rm = TRUE),
          max(our_data$radius, na.rm = TRUE),
          length.out = 400)

# build newdata matching model terms
nd <- data.frame(
  radius = xx,
  radius_squared = xx^2   # required because your model uses this column
)

# get fitted curve and draw it
yy <- predict(fit_mul, newdata = nd)
lines(xx, yy, lwd = 2, col = "red")

```


### (vii) Compare models using Adjusted \(R^2\)

From *Elementary Statistics*: the adjusted coefficient of determination penalizes extra
predictors and is
\[
\bar{R}^2 \;=\; 1 \;-\; \frac{\text{SSE}/(n-p)}{\text{SST}/(n-1)},
\]
where \(n\) is the sample size, \(p\) is the number of estimated parameters (including the
intercept), \(\text{SSE}\) is the residual sum of squares, and \(\text{SST}\) is the total sum of squares.

We use \(\bar{R}^2\) to compare:

- **Simple model:** \(A \sim x^2\) (only `radius_squared`)

- **Multiple model:** \(A \sim x + x^2\) (`radius` and `radius_squared`)

The model with the **higher** \(\bar{R}^2\) achieves a better balance of fit and parsimony.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Fit models
fit_simple   <- lm(area ~ radius_squared, data = our_data)
fit_multiple <- lm(area ~ radius + radius_squared, data = our_data)

# Extract adjusted R^2
adjR2_simple   <- summary(fit_simple)$adj.r.squared
adjR2_multiple <- summary(fit_multiple)$adj.r.squared

c(Adjusted_R2_Simple = adjR2_simple,
  Adjusted_R2_Multiple = adjR2_multiple)

```

**Interpretation:** **Results:**

- Adjusted \(R^2\) (Simple model, \(A \sim x^2\)): **0.9996**
- Adjusted \(R^2\) (Multiple model, \(A \sim x + x^2\)): **0.9995**

**Interpretation:**

Although both models explain almost all the variation in the data, the **simple regression model** with only \(x^2\) yields a slightly higher adjusted \(R^2\) (0.9996 vs. 0.9995).  

Because adjusted \(R^2\) penalizes model complexity, this indicates that adding the linear term \(x\) does not improve model performance sufficiently to justify the extra parameter. Therefore, the **simple quadratic model** provides a more balanced fit than the multiple regression model.