---
title: ""
author: ""
date: ""
output:
  pdf_document:
    latex_engine: xelatex
always_allow_html: true
fontsize: 11pt
geometry: margin=0.5in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(patchwork)

set.seed(123)

# --- model & design ---
beta0 <- 50; beta1 <- 10; sigma <- 4
x <- seq(1, 10, length.out = 20)
n <- length(x); B <- 500
Sxx  <- sum((x - mean(x))^2)
xbar <- mean(x); x0 <- 5

# --- one simulation ---
sim_one <- function() {
  y <- beta0 + beta1*x + rnorm(n, 0, sigma)
  fit <- lm(y ~ x)
  b0 <- unname(coef(fit)[1]); b1 <- unname(coef(fit)[2])

  Eyx5 <- b0 + b1*x0

  # (c) CI for slope (sigma known)
  se_b1 <- sigma / sqrt(Sxx)
  l_b1 <- b1 - 1.96*se_b1; u_b1 <- b1 + 1.96*se_b1
  cover_b1 <- (l_b1 <= beta1 && beta1 <= u_b1)

  # (d) CI for mean at x=5 (sigma known)
  se_mu <- sigma * sqrt(1/n + (x0 - xbar)^2 / Sxx)
  l_mu <- Eyx5 - 1.96*se_mu; u_mu <- Eyx5 + 1.96*se_mu
  cover_mu <- (l_mu <= (beta0 + beta1*x0) && (beta0 + beta1*x0) <= u_mu)

  c(b0 = b0, b1 = b1, Eyx5 = Eyx5,
    cover_b1 = cover_b1, cover_mu = cover_mu)
}

# --- run B times; build tibble with explicit names ---
sim_mat <- replicate(B, sim_one())
res <- as_tibble(t(sim_mat), .name_repair = "minimal")
colnames(res) <- c("b0","b1","Eyx5","cover_b1","cover_mu")

# ensure logical
res <- res |> mutate(
  cover_b1 = as.logical(cover_b1),
  cover_mu = as.logical(cover_mu)
)

```


#### (a) histograms of slope & intercept

```{r echo=FALSE, fig.align='center', fig.height=6, fig.width=12, message=FALSE, warning=FALSE, out.width="100%"}
p_b0 <- ggplot(res, aes(b0)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = beta0, linetype = 2) +
  labs(title = expression(hat(beta)[0]), x = NULL, y = NULL) +
  theme_minimal()

p_b1 <- ggplot(res, aes(b1)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = beta1, linetype = 2) +
  labs(title = expression(hat(beta)[1]), x = NULL, y = NULL) +
  theme_minimal()

p_b0|p_b1
```

**DISCUSSION:**

The histograms display the sampling distributions of the estimated intercept (\(\hat{\beta}_0\)) and slope (\(\hat{\beta}_1\)) from 500 simulated samples generated under the model 
\(y = 50 + 10x + \varepsilon,\ \varepsilon \sim N(0,16)\), with \(n = 20\) observations.

Both distributions appear **bell-shaped and symmetric**, indicating that the least-squares estimators follow an **approximately normal distribution** when the model errors are normally distributed. The dashed vertical lines represent the true parameter values \(\beta_0 = 50\) and \(\beta_1 = 10\).

The histograms are centered around their true parameter values, confirming that the estimators are **unbiased**. The slope estimator (\(\hat{\beta}_1\)) shows a narrower spread than the intercept (\(\hat{\beta}_0\)), reflecting its smaller theoretical variance:
\[
\text{Var}(\hat{\beta}_1) = \frac{\sigma^2}{S_{xx}}, \quad 
\text{Var}(\hat{\beta}_0) = \sigma^2\!\left(\frac{1}{n} + \frac{\bar{x}^2}{S_{xx}}\right).
\]
Overall, the near-normal shape and centering of the distributions validate the theoretical properties of the OLS estimators—unbiasedness, normality, and predictable variability.

#### (b) histogram of Ehat[Y|x=5]

```{r echo=FALSE, fig.align='center', fig.height=6, fig.width=12, message=FALSE, warning=FALSE, out.width="100%"}
p_mu <- ggplot(res, aes(Eyx5)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = 100, linetype = 2) +
  labs(title = expression(hat(E)[Y~"|"~x==5]), x = NULL, y = NULL) +
  theme_minimal()

p_mu
```

**DISCUSSION:**

The histogram illustrates the sampling distribution of the estimated mean response 
\(\widehat{E}[Y|x=5]\) from 500 simulated samples. Each estimate was computed using 
\(\widehat{E}[Y|x=5] = \hat{\beta}_0 + \hat{\beta}_1(5)\), where \(\hat{\beta}_0\) and 
\(\hat{\beta}_1\) are the least-squares estimates from each regression.

The distribution appears **approximately normal, symmetric, and centered near 100**, 
which is the true mean value at \(x = 5\) given by \(E[Y|x=5] = 50 + 10(5) = 100\).  
This confirms that the estimator is **unbiased**, with an expected value equal to the 
true mean response.

The histogram also shows a **narrow spread**, indicating low sampling variability — 
as expected when \(x_0\) is close to the mean of the predictor values.  
No skewness or extreme values are evident, reinforcing the conclusion that 
\(\widehat{E}[Y|x=5]\) follows an approximately normal distribution under 
the assumptions of the simple linear regression model.


#### (c) coverage for \(\beta_1\)

```{r echo=FALSE, fig.align='center', fig.height=6, fig.width=12, message=FALSE, warning=FALSE, out.width="100%"}
mean(res$cover_b1)
sum(res$cover_b1)


```

From the 500 simulated samples, **475** of the 95% confidence intervals contained the true slope value 
\(\beta_1 = 10\), giving an empirical coverage proportion of **0.95**.

This means that exactly **95%** of the constructed intervals captured the true slope — which matches the theoretical expectation for a 95% confidence interval. Therefore, the simulation confirms that the OLS-based confidence intervals for \(\beta_1\) are correctly calibrated and perform as expected under the normal error assumption.


#### (d) coverage for E[Y|x=5]

```{r echo=FALSE, warning=FALSE}
mean(res$cover_mu)
sum(res$cover_mu)
```

From the 500 simulated samples, **476** of the 95% confidence intervals for the mean response 
\(E[Y|x=5]\) contained the true value of **100**, giving an empirical coverage proportion of **0.952**.

This means that approximately **95.2%** of the constructed intervals captured the true mean response, 
which aligns closely with the theoretical expectation of 95% coverage. 

Hence, the simulation confirms that the confidence intervals for \(E[Y|x=5]\) are well-calibrated 
and perform as expected under the normal error assumption of the simple linear regression model.






